% -*- compile-command: "pdflatex class-notes-F18.tex" -*-

\documentclass{article}

\usepackage{hyperref, amsmath, xspace}
\usepackage[normalem]{ulem}

\newsavebox{\reflectbox}

\newcommand{\etc}{\emph{etc.}\xspace}
\newcommand{\eg}{\emph{e.g.}\xspace}

\newenvironment{reflect}[1]
{
  \noindent
  \begin{lrbox}{\reflectbox}
    \begin{minipage}[t]{\textwidth}
      \textbf{#1}:
}{
    \end{minipage}
  \end{lrbox}
  \fbox{\usebox{\reflectbox}}
}

\begin{document}

\title{CSCI 150: Foundations of Computer Science \\ Fall 2018 Lecture
  Notes}

\maketitle

\section{Introduction (W 8/22)}

\subsection*{Setup}

\begin{itemize}
\item {}[Meet students before class.]
\item Have students meet each other. \eg use binary cards to
  put them in groups of 4.  Then switch groups once or twice.
  Introduce people you have already met.
\end{itemize}

\subsection*{What is Computer Science?}

This class is Foundations of Computer Science.  What is Computer
Science?  (Terrible name: first, because why do you need to have
``science'' in the title?  \eg ``food product''.)  Look at each
component separately.
\begin{itemize}
\item \textbf{What is a computer?}  Get some responses.  (Can insert
  brief history lesson: Antikythera mechamism: 150--100 BC, analog
  computer.  Many other mechanical computers through the years,
  \eg Babbage Difference Engine, Zuse Z3, Curta.  IBM started off
  making machines for processing data stored on punch cards; big
  industry by 1940s. ``Computers'' referred to people.  First
  electronic computers: WWII, early 1940s.)
\item \textbf{What do computers do?}  Lots of things.  Commonality:
  information.  Communicating, transforming, analyzing, storing.
\item \textbf{What is science?}  Get some responses.  Scientists use
  rational investigation \& analysis, mathematics, \etc to study the
  natural world.
\end{itemize}

So what do computer scientists study?  \textbf{NOT computers!}
Rather, \textbf{information} = anything that can be expressed
digitally, i.e. with numbers/symbols.  ``Computer science'' is
actually a bad name, \emph{cf.}\ ``telescope science''.  Information
structure of the universe. (A much better name is ``Informatics''
which is used \eg in Europe but hasn't caught on in the US.)

Why study it?  (Maybe get some student responses)
\begin{itemize}
\item Beautiful ideas, new ways of thinking.
\item Many applications!  Can contribute directly to human flourishing.
\item Computers are everywhere.  Understanding principles of CS =
  being an informed, engaged citizen.
\end{itemize}
Could do this without a computer, but computers are excellent enabling
tools.

\subsection*{Administrivia}

Syllabus review online:
\begin{itemize}
\item Free online textbook
\item Don't bring laptops
\item Academic integrity
\item Office hours
\item Grade breakdown, labs (location), late days
\end{itemize}

For next time:

\begin{itemize}
\item Do HW 0, ``who are you''.  (Linked from webpage.)
\end{itemize}

\newpage
\section*{LAB: Minecraft Hour of Code (Th 9/23)}

\newpage
\section{Algorithms and language (F 8/24)}

\subsection*{Setup/administrivia}

\begin{itemize}
\item Read chapter 1 of textbook for Monday.
\item Homework due Monday (Minecraft programming by hand).
  Designed to help you start developing important skills, building and
  executing mental models of what is going on in the computer, to
  predict what code will do without running it.
\end{itemize}

\subsection*{Aspects of algorithms; Minecraft debugging}

\begin{itemize}
\item Book talks about 5 aspects of programming (in the reading for the
  weekend): input, output, math, conditionals, repetition.  Where did
  we see them in Minecraft?
\item 3 kinds of errors: syntax, semantic, runtime.  Talk
  about each in context of Minecraft (lots of semantic errors; runtime
  = running into wall/falling into lava?; not many opportunities for
  syntax errors (!!!), except for something like \eg a repeat loop
  block with no blocks inside it.)
\item Fact that Minecraft doesn't have (many) syntax errors is a
  Really Big Deal.  Imagine if when learning a foreign language, every
  time you made even a small grammatical mistake the other person just
  cut you off and said ``I don't understand.''  That's what it will
  feel like learning Python.  So don't be discouraged---remember what
  you could do with Minecraft.  You'll get there with Python
  too.
\end{itemize}

\subsection*{Collatz Conjecture}

\begin{itemize}
\item Write out hailstone function. Something like this:
  \begin{quote}
    Input $n$. \\
    Repeat the following until $n = 1$: \\
    \mbox{} \quad If $n$ is even, divide $n$ by $2$. \\
    \mbox{} \quad Otherwise, multiply it by $3$ and add $1$. \\
    Output the number of repetitions needed to reach $1$.
  \end{quote}
\item Note all 5 aspects of algorithm.
\item Try it on some inputs: 4, 8, 6, 11. Draw a tree etc.  Or have
  groups of students apply it to different numbers?  [In S'17 the
  students suggested starting with $7$ so we just did that on the board.]
\item Simple algorithms can have very surprising results!  Visit
  Wikipedia page.
\end{itemize}

\textbf{Mention HW assignment}: Minecraft by hand.  Posted on website.

\newpage

\section{Python I (M 8/27)}

\subsection*{Setup/administrivia}

\begin{itemize}
\item Bring color printouts of Minecraft world.
\item For this week, read Chapter 2.
\end{itemize}

\subsection*{Minecraft pair simulation exercise}

Have the students pair up and exchange programs (which they wrote for
HW).  Give each pair/group a color printout of the minecraft world for
the assignment, to use for reference.  Give them 10--15 minutes to
simulate each others' programs and confirm whether they work properly.
(See whose is shortest? Any interesting/surprising solutions?)  Then
collect the assignments.

Why did we do this?  One of the reasons the Minecraft Hour of Code lab
was much easier than the programming we will do for the rest of the
semester we already talked about---because it allows far fewer syntax
errors.  But another, perhaps more important, reason is that it gives
you immediate, visual feedback when you run your program, such that
you can follow along and quickly figure out what's happening when
something goes wrong.  When writing a Python program we won't have
that luxury.  Being able to accurately predict what a program will do
without running it on the computer is an extremely important
skill---in fact, there is actual research that shows this skill is
highly correlated with your performance in this class!  So we will
have lots more opportunities to practice it.

\subsection*{Volume of a Box}

We're going to start writing our first Python program!  Start with
something very simple: we want the computer to calculate the volume of
a box.  We will enter the width, depth, and height, and it will
multiply them.  Steps we will need:

\begin{enumerate}
\item Print a description of the program.
\item Ask user for width.
\item Remember width.
\item Ask user for height.
\item Remember height.
\item Ask user for depth.
\item Remember depth.
\item Calculate volume.
\item Print out volume.
\end{enumerate}

This involves input, output, and math.  No conditionals or repetition
yet.  Let's start learning the Python we will need to do this.  We'll
actually write the program next class.

\begin{reflect}{Brent, F'18}
  One student said it involved repetition since we repeatedly ask the
  user to enter input!  Good point.  Clarified that when we say
  ``repetition'' we specifically mean the ability to write some
  instructions once and have the computer repeat them more than once.
\end{reflect}

\subsection*{Python I (Strings, Numbers)}

Show \url{python.org} website.  Load up PyCharm, Python 3.x.

\texttt{>>>} is the prompt.  This gives you direct access to the
interpreter.  You say commands one line at a time, and get immediate
answers.  Great for testing simple expressions/lines of code and
seeing what they do.

What kinds of things does Python give us?

\begin{itemize}
\item \textbf{Strings}!

  Chunks of text, in quotes. Just like you're reading a book:
  ``Brilliant!'' Harry said.

  Use \texttt{type} to ask what type it is.

  Both types of quotes work (single, double), must end with same you
  start with. And for really big strings, use triple quotes like
  \texttt{""" lkwjelrkjwe """}

  How can we break this? \texttt{'haven't you heard?'}  Use one to
  enclose the other.  If we need both, use \texttt{"""}, or the escape
  character \texttt{\textbackslash}: \texttt{\textbackslash ", \textbackslash t,
    \textbackslash n, \textbackslash\textbackslash}. These are special.

  The red text we see when things go wrong are examples of Syntax
  Errors.

\item \textbf{Numbers}!

  2, 5, 10, integers, \etc  Use \texttt{type}.

  Simple math with numbers like $2+5$. \texttt{"2 + 5"}? not so much.
  Great as a quick calculator. Use () to force order of operations.

  Integers are great, but what about fractions? \texttt{2.5}
  \texttt{0.95}, \texttt{3e23}.

  Use \verb|type| to see the types of different things.  Mention
  exponentiation \verb|**|. (Could possibly mention mod \verb|%|, but
  I prefer to leave it until it is needed later in the semester.)

  Note that Python will automatically convert from \verb|int| to
  \verb|float| if necessary (but not vice versa).

  \begin{reflect}{Brent}
    Possible to go on and mention some things about variables if time,
    but I pretty much never make it that far.  I prefer to leave
    plenty of time for letting them ask questions, try things out,
    etc. along the way.
  \end{reflect}
\end{itemize}

\newpage

\section{Python II, Box Volume (W 8/29)}

\subsection*{Setup/administrivia}
\begin{itemize}
\item Read Chapter 2 --- quiz next class on basic Python syntax and arithmetic.
\end{itemize}

\subsection*{Variables}

For keeping stuff around in short-term memory.
\begin{verbatim}
  x = 3
  name = "Brent"
\end{verbatim}
These are \emph{assignment} statements.  No output any more. Before,
any value returned from the interpreter to us, now it goes into
memory.  Type \texttt{x}, it spits back 3.

Draw pictures, \texttt{x} has a little box containing $3$, it's just a
name for the value $3$.

\texttt{x = 4} reassigns \texttt{x} to be a new number.  \texttt{x =
  "Brent"} works too.

\texttt{y} is a runtime error, gives us an \emph{exception}, which
means something is not right (the \emph{other} meaning of
``exceptional''). So double check your names!

\textbf{Be sure to emphasize} that the expression to the right of
the \texttt{=} is \emph{evaluated} to a value which is then assigned
to the variable.  Example:
\begin{verbatim}
x = 2
y = x + 3
x = 5
\end{verbatim}
What value will \texttt{y} have afterwards?  It will still just be
$5$.  It does not change when \texttt{x} changes later.  It just
holds a number, it does not remember where the number came from.

Some names are reserved, we'll see them as we go.
Names can't be anything, letter or underscore first, then letter,
number or underscore.

Choose descriptive variable names! a, b, c, d, e, don't really help.
Use names like \texttt{width}, \texttt{height},
\texttt{num\_cats}, \etc

\subsection*{I/O, .py files}

\verb|print|.  The interpreter is great, but we don’t want the user
having to type python code, we make it easier for them.  We do
calculations, and tell the computer to print the output to the user.

\begin{itemize}
\item \verb|print(str)| --- instructs Python to display a string to
  the user.
\item For example: \verb|print("Hello world")|

\item Put stuff in a \verb|.py| file with \verb|print| and run it.

\item \verb|input()|, lets us talk to the user and ask questions
  at the same time.  Example:
\begin{verbatim}
input("What is your name?")
\end{verbatim}
\item Note it doesn't give you back the prompt \verb|>>>|, but waits
  for you to type; whatever you type comes back as a string.

\item  To do anything with this we need to save the result in a variable:
\begin{verbatim}
name = input("What is your name?")
\end{verbatim}

\item But what if we want numbers? There are functions to convert
  strings into integers or floating point numbers:
\begin{verbatim}
int("23")
float("32.4")
\end{verbatim}

\item So to get numbers from the user, we can do something like:
\begin{verbatim}
age_st = input("What is your age?")
age = int(age_st)
\end{verbatim}

  Note we get a runtime error if the user enters something that
  doesn't look like an int.
\end{itemize}

% \subsection*{REPLs, PROGRAMS, etc.}

% Mark: Useful here to talk about the difference between the REPL loop,
% and the input and print connections to the user. Draw diagram. What we
% type in python at the \verb|>>>| is the PROGRAM, it goes to the CPU,
% and the answers are returned to the PROGRAM. What happens when the
% PROGRAM says print and input, these questions come from the CPU to the
% USER and the input from the USER goes back to the CPU, which then
% returns it to the PROGRAM.

% Brent prefers something like the following: in a PROGRAM (saved in a
% file) we have to be explicit about I/O: input only comes via
% \verb|input| and output only produced via \verb|print|.  The REPL
% handles input and ouptut for us: \verb|>>>| prompt does input and
% everything is implicitly \verb|print|ed for us.  Draw box representing
% program with inputs coming in, outputs going out via these functions.
% REPL is like a little wrapper around a smaller box where we don't have
% to worry about those things.  [A little confusing to think about what
% happens when we \eg call \verb|input| at the REPL prompt but you
% don't really need to worry about that!]

% \begin{reflect}{Brent, S'17}
% Skipped this section this time around.
% \end{reflect}

\subsection*{Our first program!}

Let's start programming.  We need a place to save our commands, as a
script, so they are all run at once.  This file is produced by a
programmer.  The user will then take this file and run the program.

Create a new project and a new \verb|box.py| file in PyCharm.  First,
put name, class, date at top, in a comment block using \verb|#######|.

Add in pseudocode, use \verb|#| to mark comments, python will ignore
these lines.

Write a program to prompt the user, compute the volume of a box, and
print the result.

Once this is working, extend it to also:
\begin{itemize}
\item Compute the surface area.  (Shows off power of saving values
  entered by user and reusing them later.)
\item Compute the length of the interior diagonal
  ($\sqrt{w^2 + h^2 + d^2}$).  (Lets us explicitly talk about how to do
  roots as a fractional power, needed for Kepler/Newton lab.)
\end{itemize}

\begin{reflect}{Brent, S'17}
  I used the box example and this lecture went very well.  Things we
  discussed/questions the students had included:

  \begin{itemize}
  \item Order of operations; we played around with putting parens in
    different places and seeing whether we still got the correct
    answer.
  \item We noted that we get a runtime error if we convert with
    \verb|int| and it does not look like an int.  I asked students
    what we would need to catch this and re-ask the user; we would
    need conditionals and repetition.  Promised we would be able to do
    this later in the course.
  \end{itemize}
\end{reflect}

\newpage

\section*{More Math and Functions (mini-lecture \@ start of lab)}

\textbf{NOTE: this is here as a separate lecture, but really needs to
  just be done as a mini-lecture at the beginning of the Kepler/Newton
  lab!}

\begin{reflect}{Brent, S' 16}
  In S'16 I had already covered a bunch of this (\verb|type()|, mod,
  div) in previous lectures, so I really only had to spend 15 mintues
  talking about converting between numeric types, and importing
  \verb|math|, style check.
\end{reflect}
\begin{reflect}{Mark, F' 16}
  I was able to cover everything in lecture except the math module and
  style check. Covered math at beginning of lab, and style check 2 hours in
  once people were starting to wrap up.
\end{reflect}

\subsection*{Setup}
\begin{itemize}
\item For next time, read chapter 4.
\end{itemize}

Talk about types again.  Review, or show again, use of \verb|type()|
to determine what python thinks the type of a literal or variable is.
For example:
\begin{verbatim}
type("hello") = str
type(3) = int
type(3.14) = float
x = 3
type(x) = int
x = 3.14
type(x) = float
\end{verbatim}

Other types of formulas.  Remember, () very important in math.
Mention exponents are with \verb|**| if not already mentioned. Recall
order of operations.

We saw last time string conversion to \verb|int| and \verb|float|: Can
also use these to go from \verb|float| to \verb|int| and back.  Also,
we can go from either to string with \eg \verb|str(3)|.

What we've been using are \emph{functions}.  Some come standard with
python, like \verb|print|, \verb|input|, \verb|type|, and type
conversion functions like \verb|int|, \verb|float|, \verb|str|.

Many more possible, and in the near future, you will write your own!
For now, if you want to use functions someone else wrote, you have to
import them.

\textbf{DO THIS AT THE TOP OF THE FILE, NOT SOMEWHERE IN THE MIDDLE!}
It helps clarify what you'll need for the code, and is good etiquette.

Our first \emph{module} today, \verb|math|:
\begin{verbatim}
import math
\end{verbatim}

Lots of nice functions in there.  See
\url{http://docs.python.org/lib/module-math.html}.  For example,
\verb|log|, \verb|arctan|, \verb|ceil|, \verb|degrees|, \verb|sqrt|
and constants like \verb|e|, \verb|pi|

Can access them with dot notation: \verb|module.function|

\verb|math.log(y)| is \emph{natural} log (base $e$). If you want
\eg base 2, do \verb|math.log(10, 2)|.

Do more examples, \eg \verb|math.sqrt(math.log(math.pi))|.  Function
composition!

\begin{verbatim}
dist = int(input("What is the distance?"))
\end{verbatim}

On to Lab 3.  You will write scripts like with box volume equation,
use this as a template for your work.

Python Style Check.  Download this and use it.

\section*{LAB: Kepler \& Newton (Th 8/30)}

\textbf{Talk to TAs before lab about common errors so they can help
  give good hints when they see them.}

\begin{reflect}{Brent, S' 16}
This lab seemed to go pretty well.  Common errors:
\begin{itemize}
\item Forgetting that \eg \verb|1/3| is 0, not $0.333\dots$ (to take
  cube root).  [This is no longer a problem with Python 3.]
\item Dividing by $1.4\times 10^n$ is not the same as multiplying by
  $1.4 \times 10^{-n}$.
\item Bad parenthesization.
\item Algebra mistakes.
\end{itemize}

\end{reflect}

\begin{reflect}{Brent, S' 17}
  I had forgotten to tell them a few minor things but cleared them up
  easily during lab.  Some interesting errors I saw this time around:
  \begin{itemize}
  \item \verb|6.67 * 10e-11|.  Hint: \verb|10e-11| and \verb|1e-11|
    are not the same.  Either \verb|6.67e-11| or
    %
    \verb|6.67 * (10 ** -11)| are both fine.
  \item As usual many students missed the ``kilo'' in ``149 million
    kilometers'' and were off by a factor of 1000.
  \end{itemize}
\end{reflect}

\begin{reflect}{Brent, F'18}
  Went pretty well.  Some errors from this time:
  \begin{itemize}
  \item They didn't know about creating project vs creating a file,
    lots of projects named \texttt{orbit\_kepler.py} (not a big deal).
  \item Converting from days to seconds but then using the old days
    variable in the formula later.
  \item \verb|(4 * pi) ** 2| vs \verb|4 * pi**2|.
  \item \verb|x / y * z| vs \verb|x / (y * z)|.
  \end{itemize}
\end{reflect}

\newpage
\section{Tracing I (F 8/31)}

\subsection*{Setup}
\begin{itemize}
\item \textbf{Quiz (Python syntax and arithmetic)}.  See
  \texttt{csci150-instructors} repository.  Go over quiz answers
  afterwards (\eg have them confer with neighbors, then have
  pairs confer with other pairs, \etc)
\item HW: tracing practice (on website)
\end{itemize}

We've already mentioned that being able to manually trace the
execution of a program and predict what it is going to do is a very
important skill.  You need to be able to do this in order to figure
out how to write a program to do what you want, or to figure out why a
program is not doing what you expect.  Today we're going to introduce
a formal way to do this for Python programs.  It won't be too hard to
start out, but will get more complicated as the semester progresses
and we add new Python features, so make sure you can do it well and
accurately!

Show example of template on board: scratch work in upper left,
variables in upper right (with types!), output on the bottom (the
relative positioning is important because we're going to add extra
stuff to this basic template later in the semester!).  Work through
simple example, \eg:
\begin{verbatim}
x: int = 3
y: int = x + 2
x = 7
print("The value of y is " + str(y) + ".")
\end{verbatim}
which leads to \bigskip

\fbox{
\begin{minipage}{0.9\textwidth}
  \begin{minipage}[t]{0.5\linewidth}
    \textbf{Scratch} \medskip

    $y = x+2 \to y = 3+2 \to y = 5$
  \end{minipage}
  \begin{minipage}[t]{0.45\linewidth}
    \textbf{Variables} \medskip

    $x: \mathrm{int}\;|$ \xout{3} 7 \\
    $y: \mathrm{int}\;|$ 5
  \end{minipage} \bigskip

  \textbf{Output}

  \texttt{The value of y is 5.}
\end{minipage}
} \bigskip

The exact contents of ``scratch'' is not important.  But the contents
of the other two sections should look exactly like that.

Have them do one on their own paper:

\begin{verbatim}
z: str = "5"
msg: str = "hello"
print("z is " + z)
z = z + "2"
y: int = int(z) + 2
print(msg + "! " + z + " " + str(y))
\end{verbatim}

Go over it somehow (compare with neighbors, perhaps have someone come
up and demonstrate on the board).

Give them 5 minutes to make up a tracing exercise and then swap.  Try
to trick your neighbor!

\newpage

\section{Booleans \& Conditionals (W 9/5)}

\subsection*{Setup}

\begin{itemize}
\item Next class: tracing quiz.
\item TA hours in the lab: Tuesday 6-10pm (Kate), Weds 7-9pm (Lauren),
  Thurs 7-9pm (Fish).  There are also 151 TAs in the lab Sunday 8-10
  (Ryan) and Monday 7-10 (Innocent): if there aren't 151 students in
  need of help they are able to help with 150 as well.

  The TAs are there to help!  Make use of them!
\end{itemize}

\subsection*{Booleans}

Today: a new Python type, {\tt bool}! (Named for George Boole, 2
November 1815 -- 8 December 1864, English mathematician \&
philosopher, published ``The Laws of Thought'' in 1854 which
introduced Boolean algebra.)

Boolean values {\tt True} and {\tt False}.  Show these at the Python
prompt.  Operators {\tt and}, {\tt or}, {\tt not}.  Draw truth tables
on the board.

How else can we get Booleans?  \textbf{Comparison operators}:
\verb|>|, \verb|<|, \verb|>=|, \verb|<=|, \verb|!=|, \verb|==|.
Demonstrate on numbers, strings, all generate Booleans.  See if they
can guess what comparisons do for strings.  Come up with a hypothesis,
test it.

\subsection*{Conditionals}

You already know about conditionals.  Here's how we do them in Python.

\begin{verbatim}
passwd = input("What is the password? ")
if (passwd == "lemur"):
    print("Here be secrets.")
\end{verbatim}

Tabbing is important!  Colon is important!  The thing after {\tt if}
can be \emph{any expression which evaluates to {\tt true} or {\tt
    false}}.

Introduce random module. ({\tt import random})
\begin{itemize}
\item {\tt random.random()} is number in $[0,1)$. Uniform.
\item {\tt random.randint(n)} is random integer between $0$ and $n$
  (inclusive).  (Aside: how can we write this in terms of {\tt
    random.random()}?)
\end{itemize}
Do coin flip.  Conditionally do something.  Needs to include
\begin{itemize}
\item {\tt else}
\item {\tt elif}
\item nested {\tt if}  (flip two coins in a row)
\end{itemize}

Start with multiply nested {\tt if} and then show them {\tt elif} for
nicer syntax.

For example:
\begin{verbatim}
import random

r = random.random()
if r < 0.1:
    print("Wow, you are unlucky!")
    print("Try again next time.")
else:
    print("Yay! You get a prize!")
    if r > 0.9:
        print("You get a car!!!11")
    else:
        if r > 0.7:
            print("You get a bike!")
        else:
            if r > 0.5:
                print("You get a book!")
            else:
                if r > 0.3:
                    print("You get a sock!")
                else:
                    print("You get a stick of gum!")
\end{verbatim}

And then rewrite it to:
\begin{verbatim}
# Here's a better way to write the same thing:

r = random.random()
if r < 0.1:
    print("Wow, you are unlucky!")
    print("Try again next time.")
else:
    print("Yay! You get a prize!")
    if r > 0.9:
        print("You get a car!!!11")
    elif r > 0.7:
        print("You get a bike!")
    elif r > 0.5:
        print("You get a book!")
    elif r > 0.3:
        print("You get a sock!")
    else:
        print("You get a stick of gum!")
\end{verbatim}

\newpage
\section*{LAB: Big Data/Heart Disease (Th 9/6)}

\begin{reflect}{Brent, S'16}
  \begin{itemize}
  \item Spend a few minutes at the beginning explaining the graphs,
    especially the Boolean ones.
  \item Make sure they don't re-save the data files from Excel---it
    messes up the formatting!  Opening in Excel to \emph{look at} the
    data files is OK, just don't save.
  \item Try to guide them to using nested ifs.
  \item Make sure they understand why it's ineffective to make a huge
    condition with a bunch of ANDs.
  \item Make sure they understand why it's ineffective to have a
    string of independent ifs, where each one sets \texttt{prediction}
    (overriding whatever \texttt{prediction} was set to before).
  \end{itemize}

  Other than those gotchas, lab seems to go OK.
\end{reflect}

\begin{reflect}{Gabe, F'16}
\begin{itemize}
\item First lab section: Seems a bit short - took most students 2 hours.
\item Second lab section: Nobody finished in less than 2.5 hours. Go figure!
\item The students found the histograms very confusing. In particular, the
axes were unclear for the string-valued elements.
\end{itemize}
\end{reflect}
\begin{reflect}{Mark, F'16}
\begin{itemize}
\item can't get 100\% on 4a, since {\tt 63, False} is found in both categories. Is possible to get 100\% on 4b and 4c.
\item the chart for chest\_pain has one of the data points swapped, second bar is "asymptomatic", should have been first.
\end{itemize}
\end{reflect}
\begin{reflect}{Brent, S'17}
This lab seemed to go very well this year.  The students were
definitely engaged and learning a lot.  A few started leaving after 2
hours but probably 2/3 of the class was still working at the 3 hour
mark (but most were just finishing or very close).
\end{reflect}

\newpage

\section{Information encoding I (F 9/7)}

\subsection*{Setup/administrivia}

\begin{itemize}
\item \textbf{Quiz}: tracing. (\texttt{csci150-instructors/quiz/quiz2-tracing.py})
\item NB: exam 1 a week from today!  In class, closed notes, closed
  computer.  Covers material through today and next class. We will
  provide a practice exam on Monday; Wednesday will be review.
  \textbf{Important}: This class is about \textbf{doing} more than
  \textbf{knowing}.  One of the \textbf{worst ways} you can study is
  to review your notes and the textbook, because it gives you a false
  sense of familiarity/security without actually giving you any
  practice actually writing code.  Much better way to study is to
  \textbf{do} things: more tracing exercises; rewrite code from class
  or lab; write some code of your own.  Review book/notes as
    needed, to answer specific questions/confusions you come across.
\item Quiz on Monday re: booleans and conditionals.
\item HW due Monday (conditional practice).
\end{itemize}

\begin{reflect}{Brent, F'18}
  Skipped this aside; there was too much other stuff going on at the
  beginning of class (quiz, handing back stuff, etc.) and I wanted to
  make sure I got through the binary stuff (which I did).
\end{reflect}

\subsection*{Aside: history of information}

How many of you have heard of Ada Lovelace?  How about Charles
Babbage?

Ada Lovelace: 1815--1852, first programmer.  Wrote
programs for Charles Babbage's Analytical Engine (never built, but
definitely works in principle).  Babbage's purpose for the AE was
limited to making tables of numbers, but Lovelace had a much more
expansive and far-seeing vision:

\begin{quote}
``[The Analytical Engine] might act upon other things besides number,
were objects found whose mutual fundamental relations could be
expressed by those of the abstract science of operations, and which
should be also susceptible of adaptations to the action of the
operating notation and mechanism of the engine...

Supposing, for instance, that the fundamental relations of pitched
sounds in the science of harmony and of musical composition were
susceptible of such expression and adaptations, the engine might
compose elaborate and scientific pieces of music of any degree of
complexity or extent.''
\end{quote}

She saw potential for computers to operate on many kinds of
information instead of just making tables of numbers, and we know that
they can indeed do just that!  In fact, she even missed one very
important development, that computers can not only \emph{operate} on
these kinds of information, but allow us to \emph{communicate} using
these kinds of information as well.  Over the next 2 classes we will
consider some of the fundamentals that make this possible.

\subsection*{Binary}

Computers represent numbers using \emph{binary} = base 2 instead of
base 10.  Recall how base 10 works: ones ($=10^0$) place, tens
($=10^1$) place, hundreds ($=10^2$) place, \dots  And ten different
digits.  For example \[ 4397_{10} = 4 \times 10^3 + \dots \] Base 2 is
the same but
\begin{itemize}
\item We use $2$ instead of $10$ (ones place, 2's place, 4's place,
  8's place\dots)
\item We use $2$ digits $(0,1)$ instead of $10$.  Called \emph{bits} =
  \emph{bi}nary dig\emph{its}.
\end{itemize}
Computers do everything in binary since (1) it is no less expressive
than \eg decimal and (2) from a physical point of view, it is easier
to design hardware that can distinguish two different states
(\eg high and low voltage) than 10.

Converting base $2$ to base $10$: have them do some
examples. \eg $10110100_2 = ?$  Then do one example together on the
board, \[ 10110100_2 = 1 \times 2^7 + 1 \times 2^5 + \dots \]

Note 1 \emph{byte} = 8 bits.  Incidentally, one \emph{kilobyte} is not
$1000$ bytes, but actually $2^{10} = 1024$ bytes!  Megabyte is
$2^{20}$ bytes, and so on.

Some questions.  Pair and work on these. [Find your bit-partner:
figure out the correct base-10 number for your partner and find them
that way!]

\begin{enumerate}
\item How many different binary numbers are there using $4$ bits?
\item How many different binary numbers using $n$ bits are there?
  (Write out 16 binary numbers using $4$ bits.)
\item What is the biggest number that can be represented using $n$
  bits?
\item Convert to binary: $103_{10}$. (Don't tell them how to do it;
  see if they can figure it out.  Write a list of powers of 2 on the
  board for them.)
\end{enumerate}

Talk about answers to first four questions.  Promise we will talk
about conversion next class.

\newpage
\section{Information encoding II (M 9/10)}

\subsection*{Setup}
\begin{itemize}
\item \textbf{Quiz}: booleans and conditionals
  (\texttt{csci150-instructors/quiz/quiz3-alt-cond-practice.py}).  Put
  up on screen (e.g. in PyCharm) and have them write answers on paper.
\item \textbf{Assign project 1}---due Friday 9/21.  You might decide
  you don't want to think about it until after the exam, and that's
  OK, but you have the opportunity to start earlier if you like.
\item \textbf{Assign practice exam}---not collected or graded, but we
  will go over it on Wednesday.  In your best interest to go through
  it and come prepared!
\end{itemize}

Remember converting $103_{10}$ to binary last class?  You probably
naturally came up with something like the standard greedy algorithm.
Go over algorithm:

\begin{itemize}
\item Find largest power of two $\leq n$, say, $2^k$.
\item As long as $n > 0$:
  \begin{itemize}
  \item If $2^k \leq n$
    \begin{itemize}
    \item subtract $2^k$ from $n$
    \item write a $1$ (in the $2^k$ place)
    \end{itemize}
  \item else
    \begin{itemize}
    \item  write a $0$ in the $2^k$ place
    \end{itemize}
  \item Decrease $k$ by $1$.
  \end{itemize}
\end{itemize}

(This has all five usual aspects of algorithms: input, output, math,
conditionals, repetition.  We can't quite write this algorithm in
Python yet because we don't know how to do repetition.  Soon!)  Do an
example together on the board, \eg $103_{10}$ or something else. I've
found it seems to help some students to show them how to write out
blanks to represent different places in the binary number, then fill
them in with 1's or 0's, like so:
\[ \underset{32}{\underline{\phantom{XX}}} \; \underset{16}{\underline{\phantom{XX}}} \; \underset{8}{\underline{\phantom{XX}}} \; \underset{4}{\underline{\phantom{XX}}} \; \underset{2}{\underline{\phantom{XX}}} \; \underset{1}{\underline{\phantom{XX}}} \]
though I emphasize that I don't require them to do this.

\subsection*{Representing integers}

Integers are represented like this inside the computer.  Show entering
binary numbers directly into Python using \verb|0b10110| notation.

Most modern computers use $64$ bits to represent an integer (\emph{how
  big is that?}); some use $32$. Show Python doesn't care, abstracted
it away.  In Python 3, everything is \texttt{int}; there is also
\texttt{long} which can be as big as you want.

What about negative numbers?  One possibility: use one bit for sign
(plus or minus).  (Turns out there is a better way, ``$2$'s
complement''; learn about it in CSO or ask if you are curious.)

\subsection*{Representing text}

Basic idea: use a different number to represent each letter.  ASCII
(American Std. Code for Info. Exch.) --- early 1960's.  Specified
$128$ different characters, each using $7$ bits.  (In many cases 1 bit
left over for parity checking or just set to $0$.)
\begin{itemize}
\item Show ASCII chart.
\item Illustrate chr and ord functions in python.
\end{itemize}
$128$ characters may have been enough for the white, American,
English-speaking men who made it up.  But it sure isn't any more.
Unicode---currently over 120,000 characters. (How many bits needed?
Often uses a more complex scheme to allow different numbers of bits,
\& extending indefinitely without changing existing.  Ask if you're
curious.)

\subsection*{Representing images}

Image = grid of colored points (``pixels'' = \emph{pic}ture
\emph{el}ements).  Each pixel = mix of red, green, blue (additive
primary colors). (Have you ever put a drop of water on your phone
screen to see the pixels magnified?  It's pretty interesting, try it
if you haven't!) Each primary color has 256 possible intensities,
from 0 (off) to 255 (as bright as possible).  So each primary = 8 bits
(1 byte), each pixel = 24 bits.  How many bits/bytes for a $500 \times
500$ image?

\begin{reflect}{Brent, F'18}
  Got too carried away talking about ASCII and Unicode, images,
  \emph{etc.} (there was also a bunch of stuff at the start of class:
  handing things back, a quiz, \emph{etc.}) so I didn't actually make
  it to talking about hexadecimal.  (Apparently neither did Lars.)  So
  I sent them an email afterwards, reproduced below.
\end{reflect}

\subsection*{Hexadecimal}

Base $16$.  Need 16 symbols: 0--9, a--f. So $a_{16} = 10_{10}$ and so
on. $\mbox{ace}_{16} = ?$

Note, we can group the binary digits into 4s.  $16 = 2^4$.  So each 4
bits corresponds to 1 hexadecimal digit.  Conversion back and forth is
super easy.  We often use hexadecimal as a more convenient way to read
and write binary.  Easier for humans to read and remember.

Show entering hexadecimal directly into Python using \verb|0xace|
notation.

Show example of RGB colors expressed in hexadecimal: show
\verb|style.css| from course website.

\begin{reflect}{Brent}
  There probably won't be time to cover anything past this point.
  Including it here just in case there's extra time.
\end{reflect}

\subsection*{Representing floating-point numbers}

Back to base $10$: what does $123.45$ mean?
\[ \dots + 4 \times 10^{-1} + 5 \times 10^{-2} \]
We can do binary ``decimals'' (``binarals''?) the same way: \[
1101.011 = 1 \times 2^3 + \dots + 1 \times 2^{-1} + 1 \times 2^{-2} =
13 \frac{3}{8} \]  Also recall ``scientific notation'', \eg $1.23
\times 10^{17}$.  This is how ``floating point'' numbers are
represented: scientific notation, but in base $2$.  For example, with
$64$ bits:
\begin{itemize}
\item 1 bit for sign ($\pm$)
\item $11$ bits for exponent (base $2$ integer)
\item $52$ bits for value
\end{itemize}
\eg $-1011011 \times 2^{-3}$.

How would you represent $0.1_{10}$?  Can't represent exactly using
base $2$ (infinite)! Show $0.1 + 0.1 + 0.1$ at python prompt.

\subsection*{Video, audio}

Can mention something about how video and audio are represented.
\begin{itemize}
\item Video: sequence of frames, each frame is an image.  Of course
  there is lots of redundancy to take advantage
  of---\emph{compression}.
\item Audio: sequence of ``samples''.
\end{itemize}

\newpage
\section{Exam 1 review (W 9/12)}

\subsection*{Setup}

\begin{itemize}
\item Note we \emph{do} still have lab today/tomorrow!  Should be a
  nice review for Exam 1.
\end{itemize}

\subsection*{Sample exam}

Go over sample exam solutions.

\newpage
\section*{LAB: This Day in History (Th 9/13)}

This can be used as a comprehensive review of everything on Exam 1.

\begin{reflect}{Brent, F'18}
  For this lab they need both mod and integer division, so we covered
  them briefly at the start of lab.  I also explicitly showed them how
  to use mod 100 to find the last two digits of a number.  That aspect
  of the lab went pretty well, most students figured out how to use
  mod appropriately.

  Common mistakes (aside from the usual confusion over the rule for
  leap year) included:
  \begin{itemize}
  \item Taking the mod of \verb|dow| too early, \emph{i.e.} writing
    additional code that added offset factors to \verb|dow| in later
    steps and not reducing it mod 7 afterwards.
  \item The usual things re conditions \emph{e.g.} writing \texttt{if month
    == 3 or 5}, or \texttt{if month = 3}, or writing a sequence of
    \verb|if| instead of \texttt{if... elif... elif}.
  \end{itemize}

  As an aside, during the lab I wondered how to write the condition
  for a leap year without using nested conditionals.  Here are four
  different Boolean expressions I came up with that all work:
\begin{verbatim}
(y % 4 == 0) and ((y % 100 == 0) == (y % 400 == 0))
(y % 4 == 0) and not (y % 100 == 0 and y % 400 != 0)
(y % 4 == 0) and (y % 100 != 0 or y % 400 == 0)
(y % 4 == 0) != ((y % 100 == 0) != (y % 400 == 0))
\end{verbatim}
  Obviously we would never expect students to come up with any of these!
\end{reflect}

\newpage
\section{Exam 1 (F 9/14)}

\begin{itemize}
\item Note, no HW this weekend.
\end{itemize}

% \newpage

% \begin{reflect}{Brent, S'18}
%   Because of pushing exam 1 back one day, I only have \emph{two}
%   lectures instead of three to prepare for Guess My Number lab.  What
%   I plan to do is introduce while loops first (1 lecture), and then a
%   basic intro to functions (1 lecture).  Functions are not actually
%   super necessary for the lab, so it's OK to cover more about
%   functions in another class after the lab.

%   In the past I have used functions while doing examples for while
%   loops (\eg testing the Collatz conjecture), but that's not
%   critical---it's easy to do the examples without using loops.  It
%   might even give nice material to motivate the process of function
%   abstraction later.
% \end{reflect}

\newpage
\section{While loops (M 9/17)}

\subsection*{Setup}
\begin{itemize}
\item Reminder: Project 1 due Friday!
\end{itemize}

\subsection*{While loops}

Introduce while loops---this is a momentous day!  Once we add while
loops you can (in theory) write \emph{any program you can imagine}.
Explain syntax.  Syntax and semantics are very much like \verb|if|
(without \verb|else|): difference is that \verb|if| executes 0 or 1
time, \verb|while| executes any number of times (0 or more).

Do an example: count from 1 to 10.  Play with boundary conditions,
order of statements inside the loop.

Point out three usual components:
\begin{itemize}
\item Declare and initialize variable(s) you need
\item Write the condition with those variables
\item Change those variables somewhere in the loop (often at the
  end)
\end{itemize}

Do another example: keep printing ``hi'' and asking ``Do you want to
stop?'' until says yes --- have students write this code together on
their computers/paper.  Then have two or three share code?  There are
two decent ways to write this --- using input variable to control loop
(but then you have to initialize it with a bogus value), or using a
boolean (sentinel).

\subsection*{Collatz}

Recall Collatz Conjecture.  Idea: write a program to test the
conjecture.  Start by writing inner loop with a particular starting
number.  Then put that in a loop to do it repeatedly for different
numbers.

\begin{verbatim}
n: int = 1

max: int = int(input("Test numbers up to: "))

while n < max:
    count: int = 0
    m: int = n
    while m != 1:
        if m % 2 == 0:
            m //= 2
        else:
            m = 3*m + 1
        count += 1

    print(str(n) + ": " + str(count))

    n += 1
\end{verbatim}

(What happens if we take out \verb|max| check and just say
\verb|while true|?  A legitimate use of an infinite loop!)

\newpage
\section{Functions I: defining and calling functions (W 9/19)}

\subsection*{Setup}
\begin{itemize}
\item Reminder: project 1 due on Friday.
\item Quiz on Friday re: tracing execution with a while loop.
\end{itemize}

\subsection*{Functions}

We have used functions in Python already, like \verb|print| and
\verb|input|.

We can define our own functions!  Same rules for names as
variables. Header line: parens mean no inputs, colon, indent.  Stuff
inside the indent (body) says what happens.

\begin{verbatim}
# A function to say hi.
def say_hi():
    print("Hello there, CSCI 150!")
    print("This is a function.")
\end{verbatim}

\textbf{They must write a comment before every function!}  Comment
should clearly explain what the function does.

Function definition makes a variable which is a function object. (Note
hex value in output! =) To call it, use parentheses.  \textbf{Every
  function call must always have parentheses after it.}

Functions can be used inside other functions.

\begin{verbatim}
def say_hi_twice():
    say_hi()
    say_hi()
\end{verbatim}

Put function definitions in a script.  Note the definition itself does
not cause the function to be executed.  Python goes through the file
from top to bottom; when it reaches a definition it simply
\emph{remembers} it for later.  This means functions must be defined
before they are used, but it is perfectly OK to call a function from
inside another function \emph{before} it is defined.

\subsection*{Why functions?}

Get them to discuss why this is a worthwhile feature.  Why define
functions?  Pair \& share.

From Downey:

\begin{itemize}
\item Creating a new function gives you an opportunity to name a group
  of statements, which makes your program easier to read and debug.
\item  Functions can make a program smaller by eliminating repetitive code.
  Later, if you make a change, you only have to make it in one place.
\item Dividing a long program into
  functions allows you to debug the parts one at a time and then
  assemble them into a working whole.
\item Well-designed functions are often useful for many programs.
  Once you write and debug one, you can reuse it.
\end{itemize}

\subsection*{Parameters and arguments}

Make C to F conversion into function, takes C as argument:
\begin{verbatim}
# Input temp_C is a temperature in degrees C; print a message with
# the temperature converted to Fahrenheit.
def C_to_F(temp_C: float):
    F: float = C * 9.0 / 5.0 + 32
    print("The temperature in degrees Fahrenheit is " + str(F) + "°.")
\end{verbatim}

Note function comment should describe the input(s), if any.

\verb|temp_C| is a variable called a \emph{parameter}.  Notice we can
specify its type after a colon.

The values given to functions are called \emph{arguments}.  To call a
function, give one argument for each parameter.  The parameter will
stand for the argument.  Do some examples.  Note you can put arbitrary
expressions for arguments.

Variables created inside functions are \textbf{local}!  Parameters are also
local. Note that \verb|F| and \verb|temp_C| are not defined afterwards.

\textbf{If time}, can go on to talk about functions with return
values.

\newpage
\section*{LAB: Guess My Number (Th 9/20)}

Have the students write both versions of the game (in the first
version, the computer picks the number; in the second, the computer
guesses).

\begin{reflect}{Brent, F'18}
  Went pretty well I think.  As usual:
  \begin{itemize}
  \item I saw quite a few students tried to use recursion to do the
    ``play again'' looping.
  \item Of course many students struggled to figure out how to have
    the computer play.
  \end{itemize}
\end{reflect}

\newpage
\section{Functions II: call stack, return values (F 9/21)}

\subsection*{Setup/administrivia}

\begin{itemize}
\item Quiz: while loops
  (\texttt{csci150-instructors/quiz/while-trace.py})
\item HW due on Monday (function \& loop reading)
\end{itemize}

\subsection*{Flow of execution \& tracing with stack diagrams}

Execution starts at the first line of a program and continues.  But
when a function is called, execution jumps to that function, and when
done picks back up where it left off.  This can get complicated, since
functions can call other functions!  In general Python has to keep
track of a ``stack'' of places to resume.  Imagine you were reading a
book but it told you to go read another book and then come back.  Then
THAT book told you to go read another book, etc.  You would keep the
old books, with bookmarks, in a pile.  When you finally finish a book,
pick up the next book from the top of the pile and continue.

Do an example. Write a function that calls \verb|C_to_F| on several
temperatures.  Then write another function which does that twice, and
so on.  Then trace execution.  Introduce an error and see the
``traceback''.  Draw a stack diagram with local variables.

\begin{reflect}{Brent, F'18}
  Actually decided that this really needs to be a formal introduction
  to tracing with the function stack.  So spent longer than I have
  before on that part.  Need to write up some better notes for that
  next time.
\end{reflect}

\subsection*{``Fruitful'' functions}

Recall function from last time:
\begin{verbatim}
# Input temp_C is a temperature in degrees C; print a message with
# the temperature converted to Fahrenheit.
def C_to_F(temp_C: float):
    F: float = C * 9.0 / 5.0 + 32
    print("The temperature in degrees Fahrenheit is " + str(F))
\end{verbatim}
This function is actually really awkward.  Takes \verb|temp_C| as a
parameter, but then \emph{prints} the result.  What if we want to
compute this but not print anything?  Make each function do ONE job.
Better:
\begin{verbatim}
# Input temp_C is a temperature in degrees C; return the same
# temperature converted to degrees F.
def C_to_F(temp_C: float) -> float:
    F: float = C * 9.0 / 5.0 + 32
    return F
\end{verbatim}
Book calls it a \emph{fruitful function}.  (As a mathematician I would
just call it a \emph{function}\dots) We can specify the type that gets
returned after \verb|->|.  Note we could also get rid of the
local variable \verb|F| and just return the whole expression; matter of
taste/experience.

\verb|return| means ``stop execution of the function right now and
return this value''.  What will this do?

\begin{verbatim}
def C_to_F(temp_C: float):
    F: float = C * 9.0 / 5.0 + 32
    return F
    print("The temperature is" + str(F))
\end{verbatim}

Answer: the \verb|print| statement never happens.

However, it's OK to have multiple return statements when they are
inside \verb|if| or \verb|while|.  Let's try a function to tell us
about the temperature (\emph{do something like this, students can come
  up with details}):
\begin{verbatim}
# Convert a temperature in degrees F into an adjective describing it.
def F_to_qualitative(temp_F: float) -> str:
    if temp_F < 32:
        return "freezing"
    elif temp_F < 50:
        return "cold"
    elif temp_F < 75:
        return "warm"
    elif temp_F < 90:
        return "hot"
    else:
        return "scorching"
\end{verbatim}

Now have the students work in pairs to write a function
\verb|discuss_weather()| which prompts for temperature in C, converts
to F, and makes an appropriate comment.

\begin{verbatim}
# Have a nice chat about the weather with the user.
def discuss_weather():
    temp_C: float = float(input("What is the temperature in degrees C? "))
    temp_F: float = C_to_F(temp_C)
    word: str = F_to_qualitative(temp_F)
    print("Hmm, it's very " + word + " today, don't you think?")
\end{verbatim}

Then go through series of variants showing how you can nest function
calls, culminating in

\begin{verbatim}
# DO NOT TRY THIS AT HOME
def discuss_weather_4():
    print("It's very " +\
     F_to_qualitative(C_to_F(int(input("What is the temperature in\
     degrees C? ")))) +\
     " today, don't you think?")
\end{verbatim}

Functions can take multiple parameters.

\begin{verbatim}
# is_divisible(x,y) tests whether x is divisible by y.
def is_divisible(x: int, y: int) -> bool:
    if (x % y == 0):
        return True
    else:
        return False
\end{verbatim}

Try some examples.  Note this is actually written in a redundant way.  We
can just return \verb|x % y == 0| directly.

\subsection*{Reflections}

\begin{reflect}{Brent, S'17}
  This time I didn't get around to showing them functions with
  multiple parameters; I'll stick that in at some point when I need
  it. Also, I \emph{still} haven't shown them the mod operator---this
  was a somewhat deliberate choice since students are often confused
  by mod, and I'd rather save it until I have more motivatation for
  introducing it; perhaps at the beginning of the Caesar lab if not
  before.
\end{reflect}

\begin{reflect}{Brent, S'18}
  Once again didn't quite get to showing functions with multiple
  parameters.  This is one of those lectures that doesn't seem
  necessary, content-wise (I had already presented functions with
  return values) but it is deceptively tricky for them to build a
  correct mental model of what goes on.  I had them work in groups to
  write code on paper several times during class, and wandered around
  answering questions and clearing up confusion; it was clear that
  quite a lot of learning was going on.
\end{reflect}

\begin{reflect}{Brent, F'18}
  This time, didn't even quite get to showing the whole example with
  nesting---that is, I did show it to them, but only very quickly
  right at the end of class.  Going over the tracing stuff took more
  time.
\end{reflect}

\newpage
\section{Functions III: Abstraction (M 9/24)}

\textbf{Brent is gone at ICFP}

\subsection*{Setup}
\begin{itemize}
\item Collect HW (while loop tracing).  Go over briefly if desired?
\item Reminder: quiz on Wednesday: functions (tracing with stack)
\end{itemize}

Redo Collatz example, take code that we wrote before and successivly
abstract out parts of the code into functions.  Show how to think
about what information is going into that code (which turns into
parameters) and what information is coming out (which turns into a
return value).  Points to emphasize:

\begin{itemize}
\item Functions are for decomposing tasks \emph{hierarchically}.
  Functions aren't for chaining: \emph{i.e.}\ a function should \emph{not}
  call ``the thing that happens next'' as the last line. Can show them
  something like this example:
\begin{verbatim}
# BAD

def f1():
    do some stuff
    f2()

def f2():
    do more stuff
    f3()

def f3():
    yet more stuff
\end{verbatim}
  Instead it should be decomposed like so:
\begin{verbatim}
# GOOD

def main():
    f1()   # first step
    f2()   # second step
    f3()   # third step

def f1():
    do some stuff
    # f1 doesn't know or care what comes next after it.
    # That's main's job.

def f2():
    do more stuff

def f3():
    yet more stuff
\end{verbatim}
\item Managing information flow is very important.  Decide which
  function ``owns'' a variable.  To get it to another function---give
  it to the function as an argument.  To get a value back---return it,
  then save it in a variable.
\end{itemize}

For example you might end up with something like this:

\begin{verbatim}
# Compute the hailstone function on n.
def hailstone(n: int) -> int:
    if n % 2 == 0:
        return (n/2)
    else:
        return (3*n + 1)

# Return the number of steps in the hailstone sequence needed to get
# from n to 1.
def collatz(n: int) -> int:
    count = 0
    while n != 1:
        n = hailstone(n)
        count += 1
    return count

# Print numbers from 1 to max along with length of their Collatz chain.
def check_collatz_up_to(max: int):
    n = 1
    while n <= max:
        print (str(n) + " " + str(collatz(n)))
        n += 1
\end{verbatim}

If time, maybe take the ``volume of a box'' program we wrote a long
time ago (in the run-up to Kepler/Newton lab) and improve it by
abstracting it into functions too.

\newpage
\section{Strings (W 9/26)}

\textbf{Brent is gone at ICFP}

\subsection*{Setup}
\begin{itemize}
\item Quiz: functions
\end{itemize}

\begin{reflect}{Brent}
  Note, this continues on Friday, so it's OK if you don't quite get
  through everything, or go fast and have time to start doing more
  stuff from the beginning of Friday etc.
\end{reflect}

We sort of know about strings, but the only thing we can do with them
so far is concatenate them with \verb|+|.

\begin{itemize}
\item Strings are \emph{sequences of characters}.

\item We can reference individual characters in a string by their
  \emph{index} using brackets: \verb|mystring[3]|.  Note that
  \emph{indices start at 0}.

\item We can calculate the length of a string with \verb|len()|.
  (Check: what is the relationship of \verb|len| to valid indices?)

\item String slices: \verb|mystr[a:b]|.  Play around a bit to figure
  out what they do.
  \begin{itemize}
  \item Get the characters starting at index $a$, up to \emph{but not
      including} index $b$.  (This seems weird---at first you will just
    have to memorize this fact.  There are good reasons for it though,
    makes a lot of string indexing more elegant.)
  \item Can leave off first, or second (or both, though that is
    useless).
  \item Negative indices count backwards from the end.
  \end{itemize}

\item We've seen \verb|+|: concatenation.
\item \verb|*| does repetition.

\item Strings have \emph{methods}, i.e. functions you can call on
  strings.  Do \verb|help(str)|, \verb|help(str.upper)| etc.
\item Point out some useful methods: \verb|upper|, \verb|count|,
  \verb|replace|, \verb|find|, \verb|isdigit|
\end{itemize}

\subsection*{While loop, function, and string practice}

(OK if this moves to Friday) First, function to ``explode'' a string.
Write it collaboratively.

\begin{verbatim}
def explode(s: str):
    i: int = 0
    while i < len(s):
        print(s[i])
        i += 1
\end{verbatim}

Even better if they make some bugs in the above, run it and see what
is wrong.

(If time, do more examples from Friday.)

\newpage
\section*{LAB: Mutation is the Word (Doublets) (Th 9/27)}

\begin{reflect}{Brent S'16}
  This lab is a lot harder than the previous labs.  Definitely worth
  warning them to start early, emphasizing that they probably won't
  finish within the alloted lab time, can come and ask questions, etc.

  One of the biggest confusions typically has to do with managing
  information flow.  That is, now that we are using functions for
  everything, variables are local, so need to pass things as
  parameters and return values back.  I keep thinking about whether
  there is a gentler way to introduce this but I'm not sure.  If we
  tried to talk about ``managing information flow'' explicitly it
  would just go over their heads until they had actually grappled with
  it in a real program.  So perhaps the shock is unavoidable.

  However, see \textbf{(CRAZY?) IDEA} in reflections above, after
  coding bat lab.  Maybe switch loops and functions, so the doublets
  lab is just writing top-level code with a bunch of loops---really
  focusing on just the loops.  Then the next lab will get them to
  think about abstracting functions from existing code.
\end{reflect}

\begin{reflect}{Brent, S'17}
  The lab went a lot better this time (though it was still hard!),
  after (1) prefacing it with the ``Guess My Number'' lab which gave
  them practice writing while loops, and (2) going over the process of
  function abstraction in class, and encouraging them to first get a
  working doublets program and THEN abstract it out into functions.
\end{reflect}

\begin{reflect}{Brent, S'18}
  Used a similar approach as last time, and it seemed to go well.
\end{reflect}

\begin{reflect}{Gabe, F'18}
  I think [the lab] would benefit from some restructuring, along the
  following lines:
  \begin{itemize}
  \item Step 1: Pseudocode.
  \item Step 2: Basic implementation. No mention of functions or error
    handling. Just get it working for the case of the perfectly
    behaved user.
  \item Step 3: Incorporate error handling.
  \item Step 4: Refactor into functions.
  \end{itemize}
  I explained it this way verbally, which was helpful I think, and
  perhaps it would be easier for them to follow if we made it
  explicit. We might also consider refactoring into functions after
  the basic implementation but before the error handling.
\end{reflect}

\newpage
\section{More string, loop, \& function practice (F 9/28)}

\textbf{Brent is gone at ICFP}

\subsection*{Setup}
\begin{itemize}
\item HW due Monday (DNA strings)
\end{itemize}

Today's focus is on getting the students to do more hands-on exercises
writing code using strings, loops, and functions.

Do version of \verb|int_input| that accepts negative numbers as well
as positive (uses indexing, slicing, \verb|or|). Maybe write a simple
version of \verb|int_input| together and then have them extend it to
work with negative numbers.  For example:

\begin{verbatim}
# Inputs:
#   - prompt: string to be used as a prompt
#
# Repeatedly prompt the user for a number using the
# given prompt, until they enter a valid integer.
def int_input(prompt: str) -> int:
    valid: bool = False
    while not valid:
        uinput: str = input(prompt)
        if uinput.isdigit() or (uinput[0] == '-' and uinput[1:].isdigit()):
            valid = True
        else:
            print "That's not a valid integer.  Try again!"
    return int(uinput)
\end{verbatim}

Some more potential practice functions.  Get them to write these on
paper with partners, then go over them (perhaps even have students
come up to type them in).

\begin{verbatim}
# Compute the sum 1 + 2 + ... + n.
def sumUpTo(n: int) -> int:
    s: int = 0
    i: int = 1
    while i <= n:
        s += i
        i += 1
    return s
\end{verbatim}

\begin{verbatim}
# Keep prompting the user for a number, and stop when they enter
# zero.  Then print the sum of all the numbers the user entered.
def sumUserNums():
    s: int = 0
    n: int = 1
    while (n != 0):
        n = int(input("Please enter a number (0 to stop): "))
        s += n
    print("The sum is: " + str(s))
\end{verbatim}

\begin{verbatim}
# Print a triangle of asterisks of height n, like
#
# *
# **
# ***
#
# etc.  (Hint: build a string for each row and then print it.)
def triangle(n: int):
    rowCount: int = 1
    while rowCount <= n:
        row: str = ''
        i: int = 1
        while i <= rowCount:
            row += '*'
            i += 1
        print(row)

        rowCount += 1
\end{verbatim}

\subsection*{DNA HW intro}

Show them HW, do a bit of it together to show them what they should
do.  Make sure they understand that they should do everything at the
Python prompt, they do not need to write a program!

\newpage
\section{Lists I (M 10/1)}

\subsection*{Setup}

\subsection*{Lists}

Lists are a lot like strings, but hold arbitrary data instead of
characters.
\begin{itemize}
\item \verb|[]| is the empty list, like \verb|""| is the empty string.
\item Can hold numbers, strings, anything you want.  Show
  \verb|[1,2,3]| syntax.  Show \verb|type| on lists.
\item Elements are ordered, in a sequence.  Indexed from zero.
\item Show \verb|list(range(n))|.
\item \verb|+|, \verb|*|, \verb|len()|, slices all work the same.
\end{itemize}

\subsection*{Mutating lists}

Strings are \emph{immutable}, lists are \emph{mutable}.

\begin{verbatim}
animals = ['cat', 'dog', 'bird']
animals[1] = 'lemur'
print(animals)
\end{verbatim}
Draw a stack diagram of the above situation.

Other list mutation methods:
\begin{itemize}
\item \verb|t.append(element)| for adding to end
\item Removal:
  \begin{itemize}
  \item \verb|t.pop(index)| by index, returns deleted elt
  \item \verb|t.remove(element)| by element
  \item \verb|del t[index]| by index, doesn't return anything BUT
	\verb|del t[slice]| works too
  \end{itemize}
\end{itemize}

\begin{reflect}{Brent, S'17}
  I pared this down even more---I didn't even show them \verb|pop|,
  \verb|remove|, or \verb|del|, just showed them that they exist and
  told them to look them up.  We then had plenty of time for writing
  some practice code.  First I had them try to write
  \verb|explode_list| with partners, then we wrote it all together.
  Then, I had them work on \verb|read_inputs|, then got a volunteer to
  come up and type while the class told them what to do.  This worked
  super well!
\end{reflect}

\begin{verbatim}
from typing import *

# Input: items (list)
# Print the items in the list, one per line.
def explode_list(items: List[str]):
    n = 0
    while n < len(items):
        print(items[n])
        n += 1
\end{verbatim}

\begin{reflect}{Brent S'16}
  This is the version from S'16:
\end{reflect}

\begin{verbatim}
# Inputs: none
# Output: a list of strings entered by the user
#
# Repeatedly prompt user for inputs until they type
# 'quit', then return a list of all inputs.
def read_inputs():
    inputs = []
    done = False
    while not done:
        word = input('Enter another input, or quit: ')
        if word.lower() == 'quit':
            done = True
        else:
            inputs.append(word)

    return inputs
\end{verbatim}

\begin{reflect}{Brent, S'17}
  And these are the versions from S'17:
\end{reflect}
\begin{verbatim}
# Inputs: none
# Output: list of strings entered by the user
#
# Repeatedly prompt the user for inputs until
# they type 'quit', then return a list of all their inputs.
def read_inputs():
    answer = input("What would you like to list? ")
    stuff = []
    while answer != "quit":
        stuff.append(answer)
        answer = input("What would you like to list? ")
    return stuff

def read_inputs2():
    answer = ''
    stuff = []
    while answer != "quit":
        answer = input("What would you like to list? ")
        if answer != 'quit':
            stuff.append(answer)
    return stuff
\end{verbatim}

\newpage
\section{Lists II (W 10/3)}

\subsection*{Setup}

\begin{itemize}
\item Quiz (strings)
\item Exam 2 (primarily covering functions, while loops, strings, and
  lists---though of course it also builds on top of concepts from Exam
  1 so you still need to know those too) will be one week from today
  (day before fall break).  Practice exam will be out on Friday;
  Monday will be review.
\end{itemize}

\subsection*{Lists and strings}

First, a small aside.  Note that strings and lists of characters are
not actually the same.

\begin{itemize}
\item To convert a string to a list, you can use either:
  \begin{itemize}
  \item the \verb|list| function (every character is its own list
    item), or
  \item the \verb|split()| method.
\begin{verbatim}
  s = "banana"
  t = s.split("n")
  u = list(s)
\end{verbatim}
    Yields \verb|t = ["ba", "a", "a"]| and
    \verb|u = ['b', 'a', 'n', 'a', 'n', 'a']|.  \verb|split()| with no
    arguments splits on whitespace.  For example,
    \[ \verb|"hello      there CSCI     150!".split()| \]
    yields
    \verb|['hello', 'there', 'CSCI', '150!']|.

  \end{itemize}
\item To convert a list to a string, use the \verb|join()| method on
  the delimiter (which can be the empty string).
\begin{verbatim}
  t = ['fun', 'sun', 'sub', 'rub']
  delimiter = " -> "
  delimiter.join(t)

  ls = ['This is a line', 'This is another line', 'A third line']
  '\n'.join(ls)  # Turn a list of lines into one big string
\end{verbatim}
  \verb|''.join(...)| is the inverse of \verb|list(...)|; for example,
  \[ \verb|''.join(list('banana')) == 'banana'|. \]
\end{itemize}

\subsection*{List appending practice}

Have students find a partner. Suppose \verb|animals| is a list of
strings, how can you append 3 copies of \verb|"gerbil"| onto the end?
Try to come up with as many different ways as you can.

Go over possibilities.  This might take a while.  Try to get a bunch
of different approaches from different groups.  Talk about
\verb|append|, list repetition operator \verb|*|, concatenation
operator \verb|+|, etc.  Possibly introduce some other methods (some
correct, some incorrect) and get them to argue about whether they will
work or not, if they don't come up with enough on their own.

\begin{reflect}{Brent, S'16}
  As an illustration, here are all the suggestions my students came up
  with:
\begin{verbatim}
animals = ['skua', 'killer whale', 'spider', 'orangutan', 'goat']

# Ways to add 3 gerbils to the end of the list

# Doesn't quite work:
# animals.append('gerbil ' * 3)

# Runtime error:
# animals.append('gerbil') * 3

# Works!  (But only assuming len(animals) was 5 to start.)
# while len(animals) < 8:
#    animals.append('gerbil')

# Runtime error (too many arguments to append):
# animals.append('gerbil', 'gerbil', 'gerbil')

# Works!
# animals.append('gerbil')
# animals.append('gerbil')
# animals.append('gerbil')

# Same as 'gerbil' * 3
# animals.append('gerbil' + 'gerbil' + 'gerbil')

# Gives the right list but doesn't change animals
# animals + ['gerbil', 'gerbil', 'gerbil']

# Works!
# animals = animals + ['gerbil', 'gerbil', 'gerbil']

# Works!
# gerbil_list = ['gerbil']
# animals = animals + gerbil_list * 3

# Almost works but not quite.
# animals.append(['gerbil'] * 3)
\end{verbatim}

  I think this is a really great exercise. It really helps develop
  their intuition and their grasp of the details.
\end{reflect}

\begin{reflect}{Brent, S'17}
  The exercise was not as useful/fun this time around.  The students
  made fewer interesting mistakes, which I suppose is a good thing.
  In retrospect maybe I should have deliberately introduced more
  incorrect versions for them to think about.
\end{reflect}

\subsection*{More list practice}

\textbf{If} there's class time left (I don't plan to have any left),
could have them practice writing one or more of these functions:

\begin{verbatim}
from typing import *

# Find the sum of a list of integers.
def listsum(nums: List[int]) -> int:
    i = 0
    s = 0
    while i < len(nums):
        s += nums[i]
        i += 1
    return s

# Find the product of a list of integers.
def listprod(nums: List[int]) -> int:
    i = 0
    p = 1
    while i < len(nums):
        p *= nums[i]
        i += 1
    return p

# Find the maximum value in a nonempty list of
# integers.
def listmax(nums: List[int]) -> int:
    m = nums[0]
    i = 1
    while i < len(nums):
        if nums[i] > m:
            m = nums[i]
        i += 1
    return m
\end{verbatim}

\newpage
\section*{LAB: List Editor}

\begin{reflect}{Brent, F'18}
  I wrote this lab fresh for this semester: we decided to spend more
  time on lists and put off for loops until after exam 2.  So we
  needed a new lab that gave them practice with lists but didn't need
  for loops.

  Unfortunately the lab was something of a disaster.  Many students
  (even students who usually do well) were very confused about what
  the lab was asking them to do, and only made it to step 3 or so (out
  of 6) by the end of the lab period.  It will need to be
  substantially rewritten and/or clarified for next time (or we need
  to come up with a different lab that serves the same purpose).  I do
  think there's a good lab hiding in there, trying to get out.
\end{reflect}

\newpage
\section{Lists III (F 10/5)}

\subsection*{Setup}

\begin{itemize}
\item Practice exam is posted on the class website.  Not required and
  not collected, but good practice---we will go over it on Monday.
  Exam is Wednesday.
\item \textbf{Quiz} (\texttt{csci150-instructors/quiz/string.py})
\end{itemize}

Have them work in pairs (randomly assigned?) to write some list
practice functions.  Pick some of the following.

\begin{verbatim}
from typing import *

# Find the sum of a list of integers.
def listsum(nums: List[int]) -> int:
    i = 0
    s = 0
    while i < len(nums):
        s += nums[i]
        i += 1
    return s

# Find the product of a list of integers.
def listprod(nums: List[int]) -> int:
    i = 0
    p = 1
    while i < len(nums):
        p *= nums[i]
        i += 1
    return p

# Find the maximum value in a nonempty list of
# integers.
def listmax(nums: List[int]) -> int:
    m = nums[0]
    i = 1
    while i < len(nums):
        if nums[i] > m:
            m = nums[i]
        i += 1
    return m

# Find the index of the given string in the list, or return -1 if the
# list does not contain the string.
def listfind(search_term: str, lst: List[str]) -> int:
    i = 0
    while i < len(lst):
        if lst[i] == search_term:
            return i
        i += 1
    return -1

# Return a new list which contains only the even numbers from the
# input.  For example, onlyevens([1,5,6,2,3,3,8]) should return [6,2,8].
def onlyevens(lst: List[int]) -> List[int]:
    result: List[int] = []
    i: int = 0

    while i < len(lst):
        if lst[i] % 2 == 0:
            result.append(lst[i])
        i += 1

    return result
\end{verbatim}

\newpage
\section{Exam 2 review (M 10/8)}

Spend a lecture going over practice exam.

\section{Exam 2 (W 10/10)}

\begin{reflect}{Brent, S'16}
  In the Fall 2015 semester, Exam 2 was on the Wednesday right before
  Fall Break.  In Spring 2016, it was on a Tuesday two weeks before
  spring break. (The ``normal'' thing would have been to have the exam
  review day on Friday, then the exam on Monday, and no lab on
  Tuesday, but I was at SIGCSE on the Friday, so I gave them the
  practice exam over the weekend while I was at SIGCSE, spent Monday
  reviewing the practice exam, and held the exam during the lab slot
  for that week.)
\end{reflect}

\begin{reflect}{Brent, S'17}
  In Spring 2017, Exam 2 was on a Wednesday a week and a half before
  spring break.  For lab the following day (Thursday) I had them start
  on their projects, \emph{i.e.} not leave lab until they had a solid
  project idea.

  As usual this exam was a doozy. Average class score on the exam was
  71, median 73.5, max 100 ($\times 2$), min 21 (ouch).
\end{reflect}

\begin{reflect}{Brent, S'18}
  Again the Wednesday 1.5 weeks before spring break; lab the same \&
  following days again was used to start on project 2.

  Average was much higher this time, 83.  I made it a bit easier by
  not requiring them to *write* any for loops (though they still had
  to read several).
\end{reflect}

\begin{reflect}{Brent, F'18}
  Again on the Wednesday right before fall break.  This time the exam
  didn't even have for loops at all, since we spent more time
  practicing with lists and put off for loops until after the exam.
  Average was back down in the 71--73 range, min 28, max 100---pretty
  much the same distribution as S'17.
\end{reflect}

\newpage
\section{For loops (M 10/15)}

\subsection*{For Loops}

Recall our \verb|explode| function from a week or two ago (or just
about every list- or string-processing function we have written ever):
\begin{verbatim}
def explode(s: str):
  i: int = 0
  while i < len(s):
    print(s[i])
    i += 1
\end{verbatim}
Recall how easy it was to make errors like starting index at the wrong
place, using \verb|<=| instead of \verb|<|, and so on.  Well, it turns
out there is a better way!
\begin{verbatim}
def explode(s: str):
  for c in s:
    print(c)
\end{verbatim}
Simply loops through each character in the string one at a time.  Each
time through the loop the next character is assigned to \verb|c|.
This is way better than using a \verb|while| loop since it rules out
errors putting the starting or ending indices at the wrong number.

This also works for lists:
\begin{verbatim}
animals: List[str] = ["cow", "chicken", "pig", "rabbit"]
for animal in animals:
  print("And on this farm he had a " + animal + "!")
\end{verbatim}

General form:
\begin{verbatim}
for <new variable> in <list expression>:
    ... do stuff with variable ...
\end{verbatim}

The loop will execute a number of times equal to the length of the
list.  Each time, \verb|<new variable>| will be assigned the next
element of the list.

Have students talk in groups to come up with questions.  Get one
question from each group.  Some of the things that came up in F'18
included:

\begin{itemize}
\item Can you have if statements etc. inside a for loop?  Answer: Of
  course.  Show an example.
\item What is actually happening under the hood to make this possible?
  Answer: it's really just a shorthand for the usual sort of while
  loop we would write.
\item What happens if you change the list while looping over it?
  (Answer: who knows, don't.)
\item Can you use the variable after the end of the loop? (Answer:
  yes; don't.)
\end{itemize}

\subsection*{For loop practice}

Write this function together:
\begin{verbatim}
# Count the number of times c occurs in s.  For example,
# count('banana', 'n') = 2
# count('banana', 'z') = 0
def count(haystack: str, needle: str):
    num_occurrences: int = 0
    for c in haystack:
        if c == needle:
            num_occurrences += 1
    return num_occurrences
\end{verbatim}

Then give them this to practice writing in groups:
\begin{verbatim}
# Return a new list with only the strings of length 5 or more.
# e.g. filter_long(['hi','there','xyz','abcdefg']) = ['there','abcdefg']

def filter_long(strs: List[str]) -> List[str]:
    result: List[str] = []
    for s in strs:
        if len(s) >= 5:
            result.append(s)
    return result
\end{verbatim}

\newpage
\section{More for loops: index ranges (W 10/17)}

\subsection*{Setup}

Finish going over \verb|filter_long| from last class.

Show them, or recall, an example like this:
\begin{verbatim}
animals = ['sheep', 'horse', 'cow', 'pig']

for a in animals:
    a += '!'
\end{verbatim}

It looks like the above code adds an exclamation point to every animal
in the list, but it doesn't.  Trace the execution on the board to see
why: the variable \verb|a| changes but not the actual strings in the list!

What if we \emph{wanted} to change the animals?  We could do it with a
while loop of course, but is there a way to do it with a for loop?

Yes, there is, but it's a bit more indirect: we have to get access to
the indices, since the only way to change an element in a list is to
say something like \verb|animals[2] = ...| We can do this using
\verb|range|.

\begin{itemize}
\item Show them \verb|range|.  Type \verb|range(4,10)| at the prompt;
  doesn't evaluate.  We can turn it into a list with \verb|list| to
  see what it is: \verb|[4, 5, 6, 7, 8, 9]|.  Just like a slice: start
  at first index, up to but not including the last.
\item If we only give one index, it's like a slice without the first
  index: start at zero, go up to but not including the last.
  e.g. show \verb|range(10)|.
\item Hence we can get all the indices of a list with
  \verb|range(len(lst))|.
\end{itemize}

Like so:
\begin{verbatim}
for i in range(len(animals)):
    animals[i] += '!'
\end{verbatim}

There are other reasons we might want to talk about the indices
explicitly:
\begin{itemize}
\item To use or remember an index for later
\item To loop over two lists at the same time
\item To compare items that are next to each other
\end{itemize}

We'll see examples of all of these.  Let's practice.  Do one together first:
\begin{verbatim}
# Return the index of the first occurrence of character target in s, or -1 if it
# is not found.
def find(s: str, target: str):
    for i in range(len(s)):
        if s[i] == target:
            return i
    return -1
\end{verbatim}
Notice how we needed the index so we could return it!  This won't
work (maybe show this one first to show why it doesn't work):
\begin{verbatim}
def find(s: str, target: str):
    for c in s:
        if c == target:
            return ???  # What do we return here?
    return -1
\end{verbatim}

Now have them do one:
\begin{verbatim}
# Return a list of all the indces where target occurs in s.
def find_all(s: str, target: str) -> List[int]:
    all_the_things: List[int] = []
    for i in range(len(s)):
        if s[i] == target:
            all_the_things.append(i)
    return all_the_things
\end{verbatim}

Another one, give them this as a challenge and let them make lots of
interesting mistakes, then go over the solution:
\begin{verbatim}
# Return true if the list is sorted from smallest to biggest.
# e.g.  is_sorted([2,5,6,9,9]) = True
#       is_sorted([2,5,3,9,9]) = False

def is_sorted(nums: List[int]) -> bool:
    for i in range(len(nums) - 1):
        if nums[i] > nums[i+1]:
            return False
    return True
\end{verbatim}

Talk about how we also need indices for this one, because otherwise we
only get access to one element at a time, can't ask about neighboring
elements.  Also note why we do \verb|range(len(nums) - 1)|, so the
\verb|nums[i+1]| doesn't throw an out-of-bounds error.

% \begin{verbatim}
% def nos(s):
%     news = ""
%     for c in s:
%         if c != "s":
%             news += c
%     return news

% def sum(t):
%     total = 0
%     for n in t:
%         total += n
%     return total

% def mean(t):
%     total = 0
%     for n in t:
%         total += n
%     return float(total) / len(t)

% def mean2(t):
%     return float(sum(t)) / len(t)

% def vacation(s):
%     for i in range(len(s) - 1):
%         if s[i] == s[i + 1]:
%             return True
%     return False

% def classify(strs):
%     for s in strs:
%         print ('*' * len(s))

% def powersof2(n):
%     the_powers = []
%     for i in range(n):
%         the_powers.append(2 ** n)
%     return the_powers

% def triangle(n):
%     total = 0
%     for k in range(1,n+1):
%         total += k
%     return total
% \end{verbatim}


% \subsection*{For loops: reading exercises}

% CLOSE COMPUTERS.
% Find bit-partners. Different one for each exercise.
% Go over each and try it.

% \begin{verbatim}
% def oogie(m: List[int]) -> int:
%     p = m[0]
%     for g in m:
%         if g > p:
%             p = g
%     return p

% def yayaya(q: List[str]) -> bool:
%     for y in range(len(q) - 1):
%         if q[y] == q[y + 1]:
%             return True
%     return False
% \end{verbatim}

% Then have them write \verb|is_sorted| function.

% Additional practice:

% \begin{verbatim}
% def pulu(r: int) -> List[int]:
%     b = []
%     for k in range(r):
%         b.append(k ** 3)
%     return b
% \end{verbatim}

% \begin{reflect}{Brent, S'17}
%   Made it through all of these (including writing \verb|is_sorted|);
%   in fact, we started by finishing \verb|find(c,s)| from last class
%   (which they hadn't quite gotten to work yet).  I think it was
%   a really useful exercise.
% \end{reflect}

% \begin{reflect}{Brent, S'18}
%   Didn't make it through as many due to some technical hiccups, but
%   had them read oogie and yayaya, then had them write
%   \verb|is_sorted|.  Will need to spend a bit of time going over
%   \verb|is_sorted| next class.
% \end{reflect}

\section*{LAB: Caesar's Secrets (Th 10/18)}

\begin{reflect}{Brent, S'17}
  This lab seemed to go well, though it's definitely hard.  Most
  students were not finished by the end of the lab period.
\end{reflect}

\begin{reflect}{Brent, S'18}
  Only had time to introduce for loops at the very end of the lecture
  right before the lab.  So spent some time at the beginning of the
  lab doing more exercises writing for loops, with mixed success (I
  think it went better in Mark's lab than in mine).  In the end most
  students were making progress on the lab by the end; a few had
  finished; a few were just getting encrypt/decrypt to work.
\end{reflect}

\newpage

\section{Mutability, objects, and references (F 10/19)}

Warm-up: what happens here?  Find partner \& decide.
\begin{verbatim}
nums  = [1,2,3]
nums2 = nums
nums[1] = 0
print nums2
\end{verbatim}

Try it at python prompt.  This is really weird!

As hinted before, the fact that lists are mutable (as opposed to
strings, which are immutable) opens a huge can of worms.  Let's start
by pulling back the curtain a little bit.  Up until now we have talked
about variables having values, but we need to be more precise.  Draw
stack diagram with variables referencing two different lists.

\begin{itemize}
\item Variables actually don't contain values, they contain
  \emph{references} to objects.  A reference ``points'' to a certain
  object (really, a reference is just a number which represents a
  location in the computer's memory).
\item Copying the value of a variable into another variable just
  copies the \emph{reference} (\emph{i.e.} the memory address), so they now
  point to the same object!
\item Variables which reference objects that have the same value are
  \emph{equal}.  You can test equality with \verb|==|.
\item Variables which reference the \emph{same object} are
  \emph{identical}. You can test identity with \verb|is| (though you
  won't need this in the rest of the course).
\item Note since strings, ints, floats, and booleans are immutable, it
  didn't matter before; there is no way to tell the difference between
  two variables with separate copies of a string and two variables
  referencing the same string.  So we will continue to think of
  variables as holding ints, strings, etc., not references to them
  (since it doesn't matter).  But for lists and other things to come
  it matters quite a bit.
\end{itemize}

Do this at the REPL, with accompanying pictures drawn on the board:
\begin{verbatim}
>>> a = [1,2,3]
>>> b = [1,2,3]
>>> a == b
True
>>> a is b
False
>>> del a[0]
>>> a
[2, 3]
>>> b
[1, 2, 3]
>>> a == b
False
>>> a = b
>>> a
[1, 2, 3]
>>> b
[1, 2, 3]
>>> a == b
True
>>> a is b
True
>>> del a[0]
>>> a
[2, 3]
>>> b
[2, 3]
\end{verbatim}
Note in particular that the assignment \verb|a = b| does not copy the
value of \verb|b| into \verb|a|, but just makes \verb|a|
\emph{reference the same thing as} \verb|b|.

Some related things:
\begin{itemize}
\item What if you really \emph{do} want to make a \emph{copy} of a
  list, so modifying it doesn't modify the original?  You can write
  \verb|some_list[:]|.  This is a dirty, terrible, disgusting hack.
\item Be careful about methods that modify vs functions that make a
  new thing.  \eg \verb|list.sort()| modifies the list.  But
  \verb|sorted(list)| makes a new list!  And of course
  \verb|string.upper()| makes a new string because strings are
  immutable.  Confusing, don't memorize: the point is you should be
  careful and look up help to see what things do.  Show
  \verb|help(list.sort)|, \verb|help(sorted)|, \verb|help(str.upper)|.
\end{itemize}

[INSERT OPTIONAL RANT ABOUT MUTATION HERE]

\subsection*{Tracing with stack and heap}

This has implications for function parameters as well.  Consider:

\begin{verbatim}
def animal_release(animals):
  while len(animals) > 2:
  animals.pop(0)

print(animals)
animal_release(animals)
print(animals)
\end{verbatim}

Formally, to see what's going on, we are going to enhance our tracing
diagrams one last time: we are going to add a \emph{heap}.  The
``heap'' is the area of memory where objects are stored. Variables are
stored on the stack and contain references to things in the heap.

XXX make new template

XXX make example

\newpage


XXX SEND OUT INSTRUCTIONS/EXAMPLE RE: FILE I/O (for use with P2)

% Introduce Project 2 (word games).

% \begin{reflect}{Brent, S'18}
%   Did this at the beginning of lab instead of using class time on it.
%   Handed back exams and then went straight into recursion; this means I
%   will have three lectures on recursion and then give them the Friday
%   right before spring break off (so I can work on a conference paper
%   submission which is due that day).
% \end{reflect}

% Livecode the beginning of an exemplary Doublets lab.  Things to focus
% on:
% \begin{itemize}
% \item Start with \verb|main|.  Hierarchical decomposition; write stub
%   functions.
% \item Think about what information each function needs to do its job,
%   and what information it returns.
% \item Write \verb|get_starting_word| function.
% \item Start writing \verb|get_ending_word| function, realize that they
%   are going to have a lot of similar code.
% \item Abstract common part out into a new function, \verb|get_word|,
%   that takes a parameter describing the desired word.
% \item Stub out a single round of the game, then talk about adding a
%   while loop.
% \end{itemize}

% \begin{reflect}{Brent, S'17}
%   Instead of exam wrap-up I actually covered file I/O here, instead of
%   waiting until after spring break as I did in spring 2016. I think
%   it was useful for many students for their projects, if they wanted
%   to \eg pick random words, generate files with different word
%   lists, etc.
% \end{reflect}

\newpage
\section{Dictionaries (M 10/22)}

Recall lists.  Draw an alternate picture of lists as a mapping from
indices to values. For example \verb|['tapir', 'ferret', 'alligator']|
can be thought of as
\begin{verbatim}
0 -> 'tapir'
1 -> 'ferret'
2 -> 'alligator'
\end{verbatim}
$0$, $1$, $2$ are \emph{keys}, animal names are \emph{values}.  A
\emph{mapping} or \emph{association} between keys and values.  A
\emph{dictionary} is the obvious generalization of this to arbitrary
keys instead of just 0, 1, 2 \dots.  For example, suppose we are
keeping these animals in a zoo and we want to keep track of them by
their habitat ID numbers.
\begin{verbatim}
'A23' -> 'tapir'
'B19' -> 'ferret'
'B12' -> 'alligator'
\end{verbatim}
Or we could use numbers as keys.  Important point: keys have to be
immutable.  So we can't use \eg lists or dictionaries as keys.

Show how to make this dictionary.
\begin{itemize}
\item Write using curly braces and colons.
\item Alternatively, start with empty one and add them.
\end{itemize}

Dictionaries are like functions mapping keys to values.  Consequences:
\begin{itemize}
\item There can only be one value per unique key.  (If you want
  multiple values, use a list.)
\item But there can be multiple keys with the same value.
\item You can only look up a value by key, not a key by value.
\end{itemize}

Other things we can do with a dictionary:
\begin{itemize}
\item Get list of all keys with \verb|keys()|
\item Get list of all values with \verb|values()|
\item Check if a key is in the dictionary using \verb|in|
\item Ask for the number of keys with \verb|len|
\item Delete a mapping with \verb|del|
\item Iterate over keys with \verb|for|
\end{itemize}

Show
\url{https://docs.python.org/2.7/library/stdtypes.html#mapping-types-dict}.

\begin{reflect}{Brent, S'17}
  Only made it to here.  Did frequency counting the following class.
\end{reflect}

Rewrite \verb|frequency_counts(s)| to return a dictionary.  Why this is
better: directly look up by letter value.  No need to use \verb|ord|
or alphabet string etc.

\newpage
\begin{verbatim}
# Input: ...
# Output: ...
def frequency_counts(s):
    freqs = {}
    for c in s:
        if c not in freqs:
            freqs[c] = 0
        freqs[c] += 1

    normalize(freqs)
    return freqs

# Input: ...
# Output: ...
def normalize(d):
    total = 0
    for k in d:
        total += d[k]

    for k in d:
        d[k] /= float(total)
\end{verbatim}

\begin{reflect}{Brent, S'18}
  Made it through writing \verb|frequency_count| (but not
  \texttt{normalize}).
\end{reflect}

\newpage
\section{File I/O, dictionary practice (W 10/24)}

Our programs have no state between each time they are called.
Everything is stored in memory during the program, but afterwards,
it's all gone.  This is RAM.  We need hard drive storage to keep it
around.  Save and read things in files.

To write to a file:

\begin{verbatim}
fout = open("myfile.txt", "w")
\end{verbatim}

then to save data in the file, write out as strings:

\begin{verbatim}
fout.write("hello\n")
fout.write("world\n")
fout.close()
\end{verbatim}
Note the newline characters.

To read from a file:

\begin{verbatim}
fin = open("myfile.txt", "r")
\end{verbatim}

Two ways to get stuff out of the file.

\begin{verbatim}
s = fin.read()
\end{verbatim}

which returns a string that includes all the carriage returns, or

\begin{verbatim}
s = ""
for line in fin.readlines():
    s += line.strip()
\end{verbatim}

which strips them out of each line as it is added. \verb|readlines()|
returns a list of strings.

After both, say \verb|fin.close()| to finish accessing the file.

\subsection*{Dictionary practice}

FINISH writing \verb|frequency_count| and \verb|normalize| from last
class.  Maybe run them on the contents of a file.

\newpage
\section*{LAB: Sentiment Analysis (Th 10/25)}

\begin{reflect}{Brent, S'16}
  I wrote this lab fresh in S'16, based on a Nifty Assignment @
  SIGCSE.  It replaced the Gibberish Generator lab from the previous
  semester, which was way too difficult.  Overall I think the new lab
  went over very well.  It seemed to have the right level of
  difficulty and gave the students good practice with dictionaries.
\end{reflect}

\newpage
\section{Intro to classes \& objects (M 10/29)}

\begin{reflect}{Brent}
  Totally OK if this doesn't all fit into one lecture (it probably
  won't).  Basically we have 3 lectures to get through \verb|Cake| and
  \verb|TrafficLight|.
\end{reflect}

Birthday cake problem:
\begin{verbatim}
  # If n lit candles, blow out a random # between
  #   1 and n
  # How many blows does it take (on average)
  #   before they are all out?
\end{verbatim}

\begin{itemize}
\item Introduce idea of objects:
\begin{verbatim}
# We can package up data into "objects"
# Objects have functions, i.e. things they
#   can do. aka "Methods".

# Python "classes" allow us to design our
#   own new kinds of objects.
\end{verbatim}
\item Introduce idea of classes:
\begin{verbatim}
# class = template for objects
# an object is an "instance of" a class
# e.g. class = Car, object = my car
\end{verbatim}
\item Make empty \verb|Cake| class.
\item Show how we can make a \verb|Cake| object.
\item Show how we can set a variable, \verb|num_candles|, inside it.
  Draw stack/memory diagram.
\item Create an \verb|__init__| method with a parameter. (Explain
  \verb|self|; explain \verb|__init__|.)
\item Create \verb|blowout| and \verb|allout| methods.
\item Fill out the rest of the simulation:
\begin{verbatim}
class Cake:

    # __init__ is a special method
    #   that gets called when a
    #   new object is created.
    #   i.e. when Cake() is called.
    def __init__(self, num_candles):
        # self is a special first parameter
        # that gets filled in by Python

        # self.candles is a variable in the
        #   object being created
        #   (candles = ... would make a local var)
        self.candles = num_candles

    # Methods = things that all Cakes can do

    def blowout(self):
        n = random.randint(1, self.candles)
        self.candles -= n

    def allout(self):
        return (self.candles == 0)


def happy_birthday(c):
    count = 0
    while not c.allout():
        c.blowout()
        count += 1

    return count

def average_blows(num_candles, trials):
    total = 0
    for i in range(trials):
        c = Cake(num_candles)
        total += happy_birthday(c)
    return total / float(trials)

def main():
    for num_candles in range(100):
        print num_candles, average_blows(num_candles, 10000)

main()
\end{verbatim}
\item Run it and paste output into Excel, do a regression to show that
  it is close to natural log!
\end{itemize}

\newpage
\section{More classes \& objects}

Finish \verb|Cake| from last class. Then start on traffic light
example.  Main point: encapsulation, abstraction.

\begin{verbatim}
# Traffic light
#
# How would you model it in Python?
# What functions should it have?
# How would it work?

# Possible functions:
#   - separate red, green, yellow light functions
#   - interval function
#   - sensor
#   - change to next color
#   - say what the current color is
#   - run
# Variables / state:
#   - current color
#   - timer (time left until next change?)


# Version 1:
#   - easy to write
#   - tedious (imagine cycling through 20 colors)
#   - difficult to change
class TrafficLight:

    # Variables:
    #   - current_color (string)

    # Create a new red traffic light
    def __init__(self):
        self.current_color = "RED"

    # Change to next color
    def change(self):
        if self.current_color == "RED":
            self.current_color = "GREEN"
        elif self.current_color == "GREEN":
            self.current_color = "YELLOW"
        elif self.current_color == "YELLOW":
            self.current_color = "ORANGE"
        elif self.current_color == "ORANGE":
            self.current_color = "RED"
        else:
            print "The sky is falling!!"  # This should never happen

    # Return the current color
    def color(self):
        return self.current_color

# Version 2:
# Use a dictionary of colors
class TrafficLight2:

    # Variables:
    #   - current_color
    #   - color_dict

    def __init__(self):
        self.current_color = "RED"
        self.color_dict = \
          { "RED" : "GREEN",
            "GREEN" : "YELLOW",
            "YELLOW" : "ORANGE",
            "ORANGE" : "RED" }

    def change(self):
        self.current_color = self.color_dict[self.current_color]

    def color(self):
        return self.current_color

# Version 3: list of colors
class TrafficLight3:

    # Variables:
    #   - color_list
    #   - current_index

    def __init__(self):
        self.color_list = ["RED", "GREEN", "YELLOW", "ORANGE"]
        self.current_index = 0

    def change(self):
        self.current_index += 1
        self.current_index %= len(self.color_list)

    def color(self):
        return self.color_list[self.current_index]

def main():
    t = TrafficLight3()  # we can switch and it works the same
    print t.color()
    for i in range(10):
        t.change()
        print t.color()

main()
\end{verbatim}

\newpage
\section*{LAB: Die Hard/Water Jug (Th 11/1)}

\begin{reflect}{Brent, S'17}
  This lab seemed to go pretty well this year.  The new step that
  forced them to write a test function helped a LOT!  I know of
  several groups that started out with a buggy implementation of the
  WaterJug class and then ended up debugging it through writing tests
  and getting the tests to pass---precisely what we would hope for
  such an exercise.  We should probably figure out ways of
  incorporating this approach into other labs as well.
\end{reflect}

\newpage
\section{Classes \& objects III (F 11/2)}

Finish \verb|TrafficLight|.


\newpage
\section{Recursion I: intro (M 10/15)}

\subsection*{Setup}
\begin{itemize}
\item Reminder: project 2 due Friday!
\item Write stone-counting rules somewhere, e.g. on board (see below)
\end{itemize}

\subsection*{Stone counting}

Rules:
\begin{enumerate}
\item You may not look into the bowl.
\item You may only touch one stone.
\item If someone gives you the bowl and asks you how many stones it
  contains, you must eventually give the bowl back and answer
  truthfully.
\end{enumerate}
(There will usually be some attempts to find loopholes; modify the
rules as necessary.)  Then hand the bowl to a student and ask how many
stones it contains.

Recursion!  Functions that call themselves.  In a sense you were all
``the same function'' because you were running the same program
(rules).

Factorial function: $n! = 1 \times 2 \times \dots \times (n-1) \times
n$.  Let's implement in Python.
\begin{verbatim}
def fact(n):
    product = 1
    for k in range(1,n+1):
        product *= k
    return product
\end{verbatim}
Now consider $n! = (1 \times 2 \times \dots \times (n-1)) \times n =
(n-1)! \times n$.  Also need to define $0! = 1$.  Can we translate
this directly into Python?
\begin{verbatim}
def fact_rec(n):
    if n == 0:
        return 1
    else:
        result = fact_rec(n-1)
        return result * n
\end{verbatim}
It works!

Show call stack.  Every recursive call has its own copy of local
variables!

Big idea: transform a problem into a slightly simpler version (or more
than one), solve that simpler problem, do some work to compute an
answer to the original problem.

Typically:
\begin{itemize}
\item Base case(s): simple case where we know the answer without doing
  any work.  Always do this first!
\item Recursive case(s): make a recursive call to solve a slightly
  simpler (i.e. closer to the base case) problem.
\end{itemize}

Then re-code factorial function live, adding a bunch of comments and
  narrating thought process.

\newpage
\section{Recursion II: recursive functions (W 10/17)}

Start by reinforcing ``base case + recursive case''.  Talk about
``leap of faith'': trust that recursive calls will work without
thinking about how.  Trust your future self.  DON'T think about
unwinding entire recursion stack.  Just think about ONE level.

Up until now all the examples we've done have been math functions.
Let's do a few more ``computer-sciency'' examples processing strings
and lists.  Note, in Python this is not the best way to implement
these functions (though in some languages it is).  Just for
illustration purposes.  If you take 151 you will start to see real,
legitimate uses of recursion, \eg processing trees.  (The one
legitimate use of recursion using only what you know is what we will
be doing in lab---making pretty pictures!)

\begin{itemize}
\item Work in pairs to implement \verb|product|.
\item Have them work in pairs to implement string \verb|reverse|.
\item Have them work in (different?) pairs to implement \verb|is_palindrome|.
\end{itemize}

\begin{reflect}{Brent, S'17}
  Talked about thought process for writing recursive functions,
  re-coded factorial function.  Lots of great questions.  Then looked
  at \texttt{sum} function again which we actually wrote in the
  previous class (in retrospect it was probably premature to do
  that).  Then only got as far as having them work on writing a
  \texttt{product} function on paper.  Ended with a spirited debate
  about what \texttt{product([])} ought to be.  Plan to pick up where
  we left off next time.
\end{reflect}

\newpage
\section*{LAB: Fractal Recursion (Th 10/18)}

\begin{reflect}{S' 17}
  This lab is fun and seemed to go reasonably well, though attendance
  was poor due to being right before spring break and having project 2
  due the same day.  In the end I'm also not sure how much it really
  teaches them about recursion.  But to be honest I don't know how to
  incorporate recursion more fully into the curriculum without
  changing the programming language we use.
\end{reflect}

\newpage
\section{Recursion III (F 10/19)}

\subsection*{Setup}
\begin{itemize}
\item Project 2 due today!
\end{itemize}

\begin{reflect}{Brent, S'17}
  This semester I decided to spend three lectures on recursion instead
  of two.  It fit the schedule better (I didn't want to start a brand
  new topic for one day before spring break) and I think it is really
  needed.
\end{reflect}

Finish discussion of recursive \texttt{product} implementation,
beginning with why \texttt{product([])} should be $1$ rather than $0$:

\begin{itemize}
\item Economy of expression---more elegant definitions, with fewer
  special cases, usually means you are on the right track in
  mathematics; those definitions end up being not just more economical
  but also more useful and more consistent with other things.
\item This is the same reason $0! = 1$.
\item Validates the identity \texttt{product(xs + ys) = product(xs) * product(ys)}.
\end{itemize}

Have them work in pairs to implement string \texttt{reverse}.


\newpage
\section{Extended example I (M 11/5)}

\subsection*{ASSIGN FINAL PROJECT}

XXX write me

\subsection*{Blackjack}

As an extended example we are going to develop a simple implementation
of Blackjack.  (XXX use a different example instead?)

Card and Deck examples.
\begin{itemize}
\item Make \verb|Card| class with \verb|rank|, \verb|suit| (strings),
  and \verb|face_up| (boolean).  Though can leave \verb|face_up| until
  later.  Just has getter methods.  Write \verb|__repr__|. (If usung
  \verb|face_up|, \verb|__repr__| should return \verb|**| or something
  like that if face down.)
\end{itemize}

\newpage
\section{Extended example II (W 11/7)}

\begin{itemize}
\item Make \verb|Deck| class with list of cards.  Make \verb|__init__|
  (create 52-card deck), \verb|__repr__|, \verb|shuffle| (use
  \verb|random.shuffle|), \verb|draw| (draw one card), \verb|add|,
  \verb|deal(n)| (draw $n$ cards).  Make \verb|sort| method; go back
  and add \verb|__lt__| to \verb|Card| class.
\item If time, introduce \verb|Hand| which inherits from \verb|Deck|.
  Is-a.  Add \verb|reveal|, \verb|value| methods.
\end{itemize}

\begin{reflect}{Brent, S'17}
  I deemphasized teaching them about inheritance, instead choosing to
  just focus on encapsulation and data abstraction.  So I modified the
  example to do without inheritance.  We just used Deck for both decks
  and hands, and our Player class has a boolean to tell which kind of
  player it is.
\end{reflect}

\newpage
\section*{LAB: Processing (Th 11/8)}

\begin{reflect}{Brent, S'17}
  \textbf{VERY IMPORTANT, FOR NEXT TIME}: download Processing ahead of
  time and distribute it to students on a few USB drives!  It took
  \emph{forever} having all the students download it simultaneously at
  the start of lab.  Some students weren't really able to get started
  until over an hour into the lab period.

  Fortunately other than that the lab seemed to go well.
\end{reflect}

\newpage
\section{Extended example III (F 11/9)}

\begin{itemize}
\item Finish \verb|Hand|.
\item Introduce \verb|Player|, \verb|HumanPlayer|,
  \verb|ComputerPlayer|.
\item Finish writing blackjack.
\end{itemize}

\begin{reflect}{Brent, S'16}
  I spent 4 lectures going through blackjack in S'16.  3.5 would
  probably be enough but I couldn't quite squeeze it into 3.  I spent
  the extra half-lecture doing an empirical exploration of different
  play strategies, pitting computer players using different algorithms
  against each other in 1000's of games and printing the winning
  percentages.  Conclusions: only choosing to hit when you have a very
  small number of points (\eg 5)---or even \emph{never} choosing to
  hit---actually does much better than one might expect.  You can
  still end up winning around 40\% of the time.  On the other hand,
  still choosing to hit even when you have 17 or more points quickly
  gets very bad, because you are almost guaranteed to bust.
\end{reflect}

\begin{reflect}{Brent, S'17}
  This semester I spent 4 lectures on blackjack and we didn't even
  make it to the end!
\end{reflect}

\newpage
\section{Extended example IV (M 11/12)}

\newpage
\section{Exam 3 review (W 11/16)}

\newpage
\section{LAB: Stuckness \& Debugging (Th 11/15)}

\newpage
\section{Exam 3 (F 11/16)}

\newpage
\section{??? (M 11/19)}

\newpage
\section{THANKSGIVING}

\newpage
\section{Queues (2 lectures) (M 11/26, W 11/28)}

Start with a discussion of the queue data structure itself and why it
would be useful: model lines of people in the supermarket or at a bank
(determine why each has a different way of organizing people (multiple
lines with one checkout person each, vs one line with multiple
tellers)?), processes waiting for a turn to use the processor, network
packets waiting to be sent, etc.

Walk through the methods of a queue
(\verb|add, remove, size, is_empty|) and some important words (front,
back) and write up the super class \verb|Queue| with all the method bodies
being \verb|pass|.  We are going to write two different implementations.

Easiest way (in terms of lines of code) is to implement a queue is to
use lists: they already do everything we need, and we're just
constraining the interactions to be with the front \verb|[0]| and back
\verb|[-1]|. Code up the \verb|ListQueue| class together.

Then we test \verb|ListQueue| with a \verb|main| method (see
\verb|queuetester.py|):

\begin{verbatim}
import time

def main():
    q = ListQueue()

    for power in range(1,7):

        # Get starting time
        start = time.time()
        # Compute 10^power
        n = 10 ** power

        # Add and remove 10^power times
        for i in range(n):
            q.add(i)
        for i in range(n):
            q.remove()

        # Get ending time
        end = time.time()

        # Print power and elapsed time in seconds
        print power, end - start

main()
\end{verbatim}

As we put stress on the queue with larger numbers, we see it's not a
linear relationship, but quadratic. $10^7$ is a good place to stop,
since it takes (on the order of) 100s of seconds to complete.  Could
make a log-log plot of the size vs the time.  Give a hand-wavy
explanation why the performance is quadratic: $1 + 2 + \dots + n$.

Perhaps we could do better with a more specialized implementation,
since we only really care about the front and back.

Discuss the \verb|Node| abstraction. Break up a list into individual
pieces that fit together like Legos. It's a recursive data type,
something that stores an element and another \verb|Node| as
components.  This requires a discussion of the word \verb|None|. We've
seen it as the return value of a void function. Python has a word for
``nothing'', and we can use that word when we need a variable but it has
no value.

How do we use this to make a Queue? Make the \verb|LinkedQueue| class,
and have two components in \verb|__init__|, \verb|front| and
\verb|back|, which start off as \verb|None|.

Write \verb|size| (one of two ways: by adding recursive method to
\verb|Node|, or iterative \verb|size| method which walks down list
until current reference is \verb|None|) and \verb|is_empty|, base it
on \verb|self.front| being empty.

First, try to complete \verb|add| and \verb|remove| the natural, but
wrong, way. A node is then someone standing in line. When people stand
in lines, they look at the person in front of them. So we start with
the \verb|is_empty| case, and set \verb|set.front = Node(e)| and
\verb|self.back = self.front|. This is ok.

We then write the case when there are already people in line. We have a new
element to add to the back, so we write

\begin{verbatim}
temp = Node(e)
temp.next = self.back
self.back = temp
\end{verbatim}
to reset the \verb|back| pointer. All looks good so far.

But we can't remove people from the queue easily. We can get the
element we need from \verb|self.front|, but who is the next
\verb|self.front|? \verb|self.front.next == None|, and this is a problem. We
could walk from the back to get the element right behind \verb|self.front|,
with complicated two-finger algorithm, and it takes a long time
(totally negating any benefit from doing something other than a
list). Boo.

So instead, we reverse the orientation of everyone in line. You stand in
line backwards and stare at the person behind you. They are your next. [If
you do this in real life, people will get a little uncomfortable. Just go
with it, computer scientists are weird sometimes because it's the right
thing to do, like starting to count at 0\dots]

Then we rewrite \verb|add| to match, and finally we can write a remove
method, to set \verb|self.front = self.front.next|.  Both are now
constant time.

How does this fare with our tests? For small numbers, a little more
expensive. The overhead of making new \verb|Node| objects all the time
is seen.

But as it gets larger, the time is linear with the size. For $10^6$
and beyond, \verb|LinkedQueue| will beat \verb|ListQueue| hands
down. (Go back and plot in Excel?)

Finish with discussing how this analysis and construction of data
structures is at the core of the science part of computer science, and will
be heavily discussed in CSCI 151 with more formal tools and for more
complicated structures, within the context of Java.

\newpage
\section{LAB: Project workshop (Th 11/29)}

No assigned lab; students can come to get help on their final
projects.

\newpage
\section{Fun with Processing (F 10/30, M 12/3?)}

\begin{reflect}{Brent, S'16, S'17}
  In both S'16 and S'17 I ended with a class or two left over.  So I
  have been doing some live coding in Processing, both as an example
  of good code organization with Processing (which helps the students
  using Processing for their final project) and just for fun.  I code
  up a physically semi-realistic simulation of balls bouncing off each
  other and line segments.
\end{reflect}

\end{document}
