% -*- compile-command: "latexmk -pdf code-of-conduct-19F.tex" -*-

\documentclass{tufte-handout}

\title{CSCI 150 Lab Code of Conduct}
\date{Fall 2019, Thursday Lab}

\begin{document}

\maketitle

\section{Respect}

\begin{itemize}
  \item Learn each others' names!

  \item Don't invade personal space.  No keyboard snatching. No erasing code
  without permission. No inappropriate language directed at someone else.

  \item Don't be an ass.
\end{itemize}

\section{Communication}

\begin{itemize}
  \item Be mature about conflict. Explain advice or critiques in a
  respectful but direct manner, and be willing to accept constructive
  criticism.

  \item Listen, don't just hear. Pay attention when your partner is speaking
  and don't interrupt.
\end{itemize}

\section{Work}

\begin{itemize}
  \item Be prepared, on time, engaged and focused.

  \item Have a growth mindset.

  \item Help each other and work together; make sure you share the work
  equally and give your partner a chance to contribute.

  \item Ask for help if you're struggling.
\end{itemize}

\section{Openness}

\begin{itemize}
  \item People are different! Don't make assumptions. Value your partner's
  differences and strengths, and acknowledge your own blind spots.

  \item Be patient.

  \item Be willing to experiment and explore different ideas or solutions.

  \item Remember that prior knowledge does not indicate learning potential.

  \item Remember that you're here to learn, not show off.
\end{itemize}

\end{document}


