\documentclass{article}\oddsidemargin -.5in\evensidemargin -.5in\textwidth 7in\topmargin -.75in\textheight 9in\pagestyle{empty}\setlength{\unitlength}{.7in}\newcommand{\rectgrid}{\begin{picture}(3.6,3.6) \thicklines \put(0,1.7){\vector(1,0){3.5}} \put(1.7,0){\vector(0,1){3.5}}\linethickness{0.075mm}\multiput(.2,.2)(0.0,0.25){13}{\line(1,0){3}}\multiput(.2,.2)(0.25,0.0){13}{\line(0,1){3}} \put(1.67,3.53){$y$} \put(3.53,1.67){$x$} \end{picture}}\usepackage{graphicx}\usepackage{amssymb}\usepackage{multicol}\newcommand{\ds}{\displaystyle}\begin{document}\noindent{\bf  CSCI 150 - Fall 2018}\vspace{.15in}{\centerline{\bf Exam Bonus Assignment}}\vspace{.2in}\noindent{\bf Directions:  Rework any problem(s) on which you lost points on the exam.  You may received up to \emph{one-third} of the points back.}\vspace{0.25in}\noindent\textbf{Rules:}\begin{itemize}\item  This assignment is due Monday, October 22, at the beginning of class.  If you wake up sick, have someone else bring it to class or slip it under the door of my office.  If you are abducted by space aliens on your way to class, please ask Captain Bxvwxcryx to sign a note for you.\item  You must work each problem completely out on paper, separate from the exam itself, \emph{and in order}.    Even if you made a small mistake at the end of a problem, you must completely redo the entire problem in order to receive the bonus points for that problem.  Yes, I know I'm mean.\item  For problems with multiple parts--i.e (a), (b), (c)--you only need to work the part(s) on which you lost points.\item  You must explain your work as you go, \emph{using complete sentences.}  Your job is to convince me that you now understand how to work the problem.\item  You must turn in your original exam along with you corrections.  Please do not write on or change your original exam in any way.\item  I will be somewhat more harsh in giving any partial credit than on the original exam itself.  However, you do not need to work all problem(s) on which you lost points.  You may do as many or few as you like.\item  You are welcome and encouraged to work together.  I will help and give hints, but will not work the problem(s) out, nor will I necessarily tell you if your answer(s) are correct.\item  I will round the total points you receive to the nearest half point.\end{itemize}\vspace{0.2in}\noindent\textbf{Rationale:}  Though the points are a nice incentive to complete the assignment, that is not the main reason I offer this.  Foundations of Computer Science is a cumulative process, and it is essential for your successful completion that you eventually master this material.  In addition, even in classes where no bonus points are offered, going over your returned exams in detail is a habit that I want to encourage.  I \emph{strongly} encourage you to start on this assignment soon--those who wait until the last moment rush through it, make mistakes (and therefore get very little credit), and most importantly still don't really understand how to work the problems.  \emph{Please do not hesitate} to get help.\end{document}