% -*- compile-command: "pdflatex exam-bonus.tex" -*-

\documentclass{article}

% \oddsidemargin -.5in
% \evensidemargin -.5in
% \textwidth 7in
% \topmargin -.75in
% \textheight 9in
\pagestyle{empty}
\setlength{\unitlength}{.7in}

\usepackage{graphicx}
\usepackage{amssymb}

\begin{document}

\noindent \textbf{CSCI 150/151, Spring 2019}
\textbf{Exam Bonus Assignment}

\bigskip

\noindent{\bf Directions: Rework any problem(s) on which you lost
  points on the exam.  You may receive up to one half of the points
  back.}

\bigskip

\noindent\textbf{Guidelines:}
\begin{itemize}
\item  This assignment is due by Friday, March 29, at the beginning of
  class, no exceptions.
\item You must work each problem completely out on paper, separate
  from the exam itself, \emph{in order}.  You must completely redo the
  entire problem in order to receive the bonus points for that
  problem.\footnote{If you are upset that you have to redo an entire
    problem just to correct one small mistake, you should consider
    whether it is really worth your time to submit a corrected version
    of that problem at all.}
\item For problems with multiple parts---\emph{i.e.} (a), (b),
  (c)---you only need to do the part(s) on which you lost points.
\item You must explain your work as you go, \emph{using complete
    sentences.}  Your job is to convince me that you now understand
  how to do the problem.
\item You must turn in your original exam along with your corrections.
  Please do not write on or change your original exam in any way.
\item You do not need to work all problem(s) on which you lost points.
  You may do as many or few as you like.
\item You are welcome and encouraged to work together.  I will help
  and give hints, but will not work the problem(s) out, nor will I
  necessarily tell you if your answer(s) are correct.
\end{itemize}

\end{document}
